---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# MicroBioMeta

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/Steph0522/MicroBioMeta)](https://github.com/Steph0522/MicroBioMeta/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/Steph0522/MicroBioMeta)](https://github.com/Steph0522/MicroBioMeta/pulls)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Bioc release status](http://www.bioconductor.org/shields/build/release/bioc/MicroBioMeta.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/MicroBioMeta)
[![Bioc devel status](http://www.bioconductor.org/shields/build/devel/bioc/MicroBioMeta.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/MicroBioMeta)
[![Bioc downloads rank](https://bioconductor.org/shields/downloads/release/MicroBioMeta.svg)](http://bioconductor.org/packages/stats/bioc/MicroBioMeta/)
[![Bioc support](https://bioconductor.org/shields/posts/MicroBioMeta.svg)](https://support.bioconductor.org/tag/MicroBioMeta)
[![Bioc history](https://bioconductor.org/shields/years-in-bioc/MicroBioMeta.svg)](https://bioconductor.org/packages/release/bioc/html/MicroBioMeta.html#since)
[![Bioc last commit](https://bioconductor.org/shields/lastcommit/devel/bioc/MicroBioMeta.svg)](http://bioconductor.org/checkResults/devel/bioc-LATEST/MicroBioMeta/)
[![Bioc dependencies](https://bioconductor.org/shields/dependencies/release/MicroBioMeta.svg)](https://bioconductor.org/packages/release/bioc/html/MicroBioMeta.html#since)
[![R-CMD-check-bioc](https://github.com/Steph0522/MicroBioMeta/actions/workflows/R-CMD-check-bioc.yaml/badge.svg)](https://github.com/Steph0522/MicroBioMeta/actions/workflows/R-CMD-check-bioc.yaml)
<!-- badges: end -->

The goal of `MicroBioMeta` is to support beginners in microbiome data analysis, whether working with metagenomic or metabarcoding datasets. It enables users with limited experience in R to explore, visualize, and analyze results with minimal coding and customization. This helps save time on scripting, while still producing statistically robust results and publication-ready figures in an intuitive and accessible way

## Contributors

- Stephanie Hereira - lead developer.
- Nina Montoya - co-developer and functions.
- Karla Zarco - co-developer and functions.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `MicroBioMeta` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("MicroBioMeta")
```

And the development version from [GitHub](https://github.com/Steph0522/MicroBioMeta) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("Steph0522/MicroBioMeta")
```
## Examples

This is a basic example which shows you how to solve a common problem:

```{r example, eval = requireNamespace('MicroBioMeta')}
library("MicroBioMeta")
## basic example code
```

## Citation

Below is the citation output from using `citation('MicroBioMeta')` in R. Please
run this yourself to check for any updates on how to cite __MicroBioMeta__.

```{r 'citation', eval = requireNamespace('MicroBioMeta')}
print(citation("MicroBioMeta"), bibtex = TRUE)
```

Please note that the `MicroBioMeta` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `MicroBioMeta` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://Steph0522.github.io/MicroBioMeta) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.


