---
title: "Metabarcoding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{metabarcoding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

For this example, we will use *metabarcoding* data, specifically 16S for bacteria and 18S for fungi.

The data is taken from:

For 16S: [Hereira-Pacheco, S.E., Navarro-Noya, Y.E. & Dendooven, L. The root endophytic bacterial community of Ricinus communis L. resembles the seeds community more than the rhizosphere bacteria independent of soil water content. Sci Rep 11, 2173 (2021). https://doi.org/10.1038/s41598-021-81551-7](https://doi.org/10.1038/s41598-021-81551-7)]

For 18S: [Hereira-Pacheco, S. E., Estrada-Torres, A., Dendooven, L., & Navarro-Noya, Y. E. (2023). Shifts in root-associated fungal communities under drought conditions in Ricinus communis. Fungal Ecology, 63, 101225.](https://www.sciencedirect.com/science/article/abs/pii/S1754504823000028)

First, Let's call the `MicroBioMeta` package:

```{r "start", message=FALSE}
library("MicroBioMeta")
```

Let's call other packages needed:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(qiime2R)
```

[`tidyverse`](Wickham,%20H.,%20Averick,%20M.,%20Bryan,%20J.,%20Chang,%20W.,%20McGowan,%20L.%20D.%20A.,%20François,%20R.,%20...%20&%20Yutani,%20H.%20(2019).%20Welcome%20to%20the%20Tidyverse.%20Journal%20of%20open%20source%20software,%204(43),%201686.) is used for all the data wrangling and visualizations and [`qiime2R`](https://github.com/jbisanz/qiime2R) for calling `.qza` objects as output from [`QIIME2`](Estaki,%20M.,%20Jiang,%20L.,%20Bokulich,%20N.%20A.,%20McDonald,%20D.,%20González,%20A.,%20Kosciolek,%20T.,%20...%20&%20Knight,%20R.%20(2020).%20QIIME%202%20enables%20comprehensive%20end‐to‐end%20analysis%20of%20diverse%20microbiome%20data%20and%20comparative%20studies%20with%20publicly%20available%20data.%20Current%20protocols%20in%20bioinformatics,%2070(1),%20e100.).

Then, Load the data and reformat. Notice the data we are loading came from processing in [`QIIME2`](Estaki,%20M.,%20Jiang,%20L.,%20Bokulich,%20N.%20A.,%20McDonald,%20D.,%20González,%20A.,%20Kosciolek,%20T.,%20...%20&%20Knight,%20R.%20(2020).%20QIIME%202%20enables%20comprehensive%20end‐to‐end%20analysis%20of%20diverse%20microbiome%20data%20and%20comparative%20studies%20with%20publicly%20available%20data.%20Current%20protocols%20in%20bioinformatics,%2070(1),%20e100.) , so the `read_qza` function is used to import this data into R, also we could export the data within [`QIIME2`](Estaki,%20M.,%20Jiang,%20L.,%20Bokulich,%20N.%20A.,%20McDonald,%20D.,%20González,%20A.,%20Kosciolek,%20T.,%20...%20&%20Knight,%20R.%20(2020).%20QIIME%202%20enables%20comprehensive%20end‐to‐end%20analysis%20of%20diverse%20microbiome%20data%20and%20comparative%20studies%20with%20publicly%20available%20data.%20Current%20protocols%20in%20bioinformatics,%2070(1),%20e100.) and then import the data as `.txt`.

```{r}

bacteria_table <- qiime2R::read_qza("../man/Data/dada2.200.trimmed.single.filtered.cleaned.qza")$data %>% as.data.frame()

fungi_table <- qiime2R::read_qza("../man/Data/merge_table_240_noplant_filtered_nous.qza")$data %>% 
  as.data.frame()

metadata <- read.delim("../man/Data/FINALMAP.txt", check.names = F) %>% rename(SAMPLEID =`#SampleID`) %>% 
  filter(Month == "2") %>%
  mutate(Treatment = factor(
    case_when(
      Treatment == 1 ~ "TC",
      Treatment == 2 ~ "TD",
      Treatment == 3 ~ "TED",
      TRUE ~ as.character(Treatment)
    ),
    levels = c("TC", "TD", "TED")
  ))

metadata2 <- read.delim("../man/Data/FINALMAP18S.txt", check.names = F)%>% rename(SAMPLEID=`#SampleID`)%>%
  mutate(
    Treatment = factor(
      case_when(
        Treatment == 1 ~ "TC",
        Treatment == 2 ~ "TD",
        Treatment == 3 ~ "TED",
        TRUE ~ as.character(Treatment)
      ),
      levels = c("TC", "TD", "TED")
    )
  )



bacteria_table_filter <- bacteria_table[match(metadata$SAMPLEID, colnames(bacteria_table)), ]


taxonomy_bacteria <- read_qza("../man/Data/taxonomy_dada2.200.trimmed.single.seqs.qza")$data %>%
  as.data.frame() %>% rename(taxonomy = Taxon) %>% 
  dplyr::select(-Confidence) %>% 
  column_to_rownames(var = "Feature.ID")

taxonomy_fungi <- read_qza("../man/Data/taxonomy_blast_240_0.97.qza")$data %>% as.data.frame() %>% 
  rename(taxonomy = Taxon) %>% 
  dplyr::select(-Consensus) %>% 
  column_to_rownames(var = "Feature.ID")


```

## Pre-processing data

In order to use the data as input for functions of `MicroBioMeta` should be in a `biom` format with the column of **taxonomy** at the end.

In case your data is not in a `biom` form, we could use the `merge_feature_taxonomy` function to set the data in this format. This format will allow the use of taxonomy identity without load another object into R.

```{r}

table_bac <- merge_feature_taxonomy(table = bacteria_table_filter,
                                    taxonomy = taxonomy_bacteria)

table_fung <- merge_feature_taxonomy(table = fungi_table, 
                                     taxonomy = taxonomy_fungi)

```

## Composition exploration

`MicroBioMeta` has several functions to explore composition, taxonomic identity, and abundance. Let's explore each of them.

### Abundance barplots

Abundance barplots are useful to identify patterns in community composition based on taxonomic identity.\
This function generates stacked barplots of relative abundances using an abundance table and a metadata file as input.

The `taxonomy_db` parameter specifies the taxonomic reference database used for taxonomic assignment (e.g., SILVA, GTDB, UNITE).\
The `x_col` parameter indicates the column in the metadata used for grouping samples along the x-axis, while `facet_col` defines the metadata column used to split the plot into facets.

The `level` parameters allow us to choose the taxonomic level to collapse.

The `top_n_groups` controls the number of most abundant taxa displayed in the barplot.

If `add_remain = TRUE`, the relative abundances are completed to 100% by adding a gray category representing the remaining taxa not explicitly displayed; if `FALSE`, only the selected taxa are shown.\
The `label` parameter defines the legend title displayed in the plot.

Finally, when `save.table = TRUE`, the function also returns a table containing the relative abundances (percentages) used to generate the plot.\

```{r}
abundance_barplot(table = table_bac,
                  metadata = metadata,
                  taxonomy_db = "silva",
                  level = "phylum", 
                  top_n_groups = 15,
                  x_col = "Type_of_soil", 
                  facet_col = "Treatment", 
                  add_remained = TRUE,
                  label =   "Phylum",
                  save_table = FALSE)+
  theme(axis.text.x = element_text(angle = 50, hjust = 0.95))

```

```{r}
abundance_barplot(table = table_fung,
                  metadata = metadata2,
                  x_col = "Treatment", 
                  facet_col = "Type_of_soil", 
                  add_remained = TRUE,  
                  label =   "Genera",
                  save_table = FALSE)+
  theme(axis.text.x = element_text(angle = 50, hjust = 0.95))
```

### Abundance heatmaps

The input `table` must be a data frame containing taxonomic features (rows) and samples (columns).\
The `metadata` argument provides sample-level information used to annotate the heatmap.

The parameters `condition1`, `condition2`, and `condition3` define up to three metadata variables to be displayed as horizontal annotations above the heatmap.\
Custom color palettes for each annotation can be specified using `colors_condition1`, `colors_condition2`, and `colors_condition3`, respectively.\
Legend titles for these annotations can be customized with `name_legend_condition1`, `name_legend_condition2`, and `name_legend_condition3`.

The `top_n` argument controls the number of most abundant features included in the heatmap.\
If `cluster = TRUE`, rows are hierarchically clustered; if `FALSE`, features are ordered by decreasing relative abundance.\
The `show_column_names` parameter determines whether sample names are displayed in the heatmap.

```{r, warning=FALSE}
abundance_heatmap_plot(table = table_bac,
                  metadata = metadata,
                  top_n = 30,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil", 
                  condition2 = "Treatment")
```

```{r, warning=FALSE}
abundance_heatmap_plot(table = table_fung,
                  metadata = metadata2,
                  top_n = 30,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil", 
                  condition2 = "Treatment")
```

## Alpha diversity

Alpha diversity describes the diversity within individual samples. In `MicroBioMeta`, alpha diversity can be explored using several metrics, including traditional indices such as richness, Shannon, and Simpson. In addition, the package implements functions based on the **Hill numbers framework**, which provides a unified way to represent these indices as members of the same family of diversity measures. Hill numbers are parameterized by the order *q*, which controls the sensitivity of the metric to species abundances. For example, *q*=0 corresponds to species richness, *q*=1 to the exponential of Shannon entropy, and *q*=2 to the inverse Simpson index. This formulation facilitates consistent comparison among diversity measures and allows researchers to evaluate how diversity patterns change depending on the weight given to rare versus abundant taxa.

### Correlation among Hill numbers

The `alpha_hill_corrplot()` function calculates Hill numbers for different orders (typically, 𝑞=0,1,2) and visualizes the correlations among them using a correlation plot.

This plot helps evaluate how strongly diversity metrics of different orders are related to the depth sequencing.

The table argument must be a data frame containing features (rows) and samples (columns).

```{r, warning=FALSE}
alpha_hill_corrplot(table = table_bac)

```

```{r, warning=FALSE}
alpha_hill_corrplot(table = table_fung)
```

### Alpha diversity visualization

The `alpha_hill_plot()` function calculates Hill numbers and visualizes alpha diversity patterns across experimental conditions using boxplots.

The `metadata` argument provides sample-level information used to group and visualize the samples.

The `x_col` parameter specifies the metadata column used on the x-axis, while `fill_col` controls the grouping used to color the boxplots.

The `facet_by` argument allows splitting the plot according to a metadata variable, and `facet_orientation` controls whether facets are arranged horizontally or vertically.

If `save_table = TRUE`, the function also returns a table containing the alpha diversity values calculated for each sample.

```{r, warning=FALSE}
alpha_hill_plot(table = table_bac, 
                metadata = metadata%>% filter(SAMPLEID %in% colnames(table_bac)),
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "vertical",
                save_table = FALSE)
```

```{r, warning=FALSE}
alpha_hill_plot(table = table_fung, 
                metadata = metadata%>% filter(SAMPLEID %in% colnames(table_fung)),
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                save_table = FALSE)
```

## Beta diversity

```{r, warning=FALSE, message=FALSE}
beta_div_plot(table = table_bac, 
              metadata = metadata, 
              group_col = "Type_of_soil",
              shape_col = "Treatment")
```

```{r, warning=FALSE, message=FALSE}
beta_div_plot(table = table_fung, 
              metadata = metadata2, 
              group_col = "Type_of_soil",
              shape_col = "Treatment",distance = "aitchison", ordination = "NMDS")
``
