---
title: "Metagenomic"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Metagenomic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

🍄 This tutorial shows how to use `MicroBioMeta` on **shotgun metagenomic** data, as an alternative to the amplicon (metabarcoding) workflow covered in the [Metabarcoding vignette](metabarcoding.html). The functions and most of their parameters are exactly the same across both workflows, so they aren't re-explained here only what's specific to metagenomic data is.

For this example, reads were taxonomically classified with [`Kraken2`](https://ccb.jhu.edu/software/kraken2/) [@wood2019kraken2] and abundance-corrected at the species level with [`Bracken`](https://ccb.jhu.edu/software/bracken/) [@lu2017bracken].

The data comes from forest soils along the Iztaccíhuatl-Popocatépetl (PNIP) and La Malinche (PNML) transects, Central Mexico:

[Hereira-Pacheco, S. E. et al. (2025). Metagenomic analysis of the diversity and composition of the fungal communities of forest soils of the transect Iztaccíhuatl-Popocatépetl and La Malinche. *PeerJ*, 13, e18323.](https://peerj.com/articles/18323/)

Raw data, metadata and the original analysis scripts: [github.com/Steph0522/Fungal_Metagenomes_PNIP_PNML](https://github.com/Steph0522/Fungal_Metagenomes_PNIP_PNML)

## 💻 Loading data
First, let's load `MicroBioMeta` and `tidyverse` (used for data wrangling):

```{r "start", message=FALSE}
library("MicroBioMeta")
library(tidyverse)
```

Each sample was processed separately, so `bracken_species` reports were merged into a single table beforehand (one column per sample, named `kraken_pluspfp_<id_metagenome>_report_bracken_species`) with a **`taxonomy`** column already attached as the last column:

```{r}
table_kraken <- read.delim("articles/data/table_kraken.txt", row.names = 1, check.names = FALSE)

table_kraken[1:3, c(1:2, ncol(table_kraken))]
```

Taxonomy is already in the table  so, there is no need to use  `merge_feature_taxonomy()` function.

⚠️ Taxonomy strings from Kraken2 use a `k__`/`p__`/`c__`/... prefix like SILVA/GTDB, but ranks are separated by `"; "` (semicolon **and space**) instead of `";"`. `MicroBioMeta` already accounts for this wherever a `taxonomy_db` argument exists, just pass `taxonomy_db = "Kraken2"`.

Now, let's load the metadata:

```{r}
metadata_meta <- read.delim("articles/data/metadata_metagenomic.txt", check.names = FALSE)

head(metadata_meta[, 1:10])
```

Kraken column names encode a `id_metagenome` number (e.g. `kraken_pluspfp_100_report_bracken_species`) rather than the sample's actual ID, so we recover it with a regular expression and use `metadata_meta$id_metagenome` to rename each column to its `SampleID`:

```{r}
sample_cols <- setdiff(colnames(table_kraken), "taxonomy")

id_metagenome <- as.integer(gsub("kraken_pluspfp_|_report_bracken_species", "", sample_cols))

colnames(table_kraken)[match(sample_cols, colnames(table_kraken))] <-
  metadata_meta$SAMPLEID[match(id_metagenome, metadata_meta$id_metagenome)]
```

Let's rename and format the metadata:

```{r}
metadata_meta$Polygon <- factor(paste0("Pol", metadata_meta$Poligono), levels = paste0("Pol", 1:6))
metadata_meta$Site <- as.character(metadata_meta$Sitio)
```

```{r}
table_meta <- table_kraken[, c(metadata_meta$SAMPLEID, "taxonomy")]
```

Coordinates were taken and loaded:

```{r}
coord_meta <- read.csv("articles/data/coord_metagenomic.csv") %>%
  dplyr::select(-Site) %>% # drop the CSV's own overall site index (1-12); we want `Sitio` (1-2) instead
  dplyr::rename(Polygon_num = pol, Site = Sitio) %>%
  dplyr::select(Polygon_num, Site, Transecto, Latitude, Longitude)

metadata_meta <- metadata_meta %>%
  dplyr::mutate(Polygon_num = as.integer(gsub("Pol", "", Polygon)), Site = as.integer(Site)) %>%
  dplyr::left_join(coord_meta, by = c("Polygon_num", "Site", "Transecto")) %>%
  dplyr::mutate(Site = as.character(Site))
```

## 📊 Composition exploration
### 📊 Abundance barplot
Same function as in the metabarcoding vignette; the only difference is `taxonomy_db = "Kraken2"`, which tells the taxonomy-parsing logic to expect Kraken2/Bracken-style strings.

```{r, fig.width=9, fig.height=6}
abundance_bar_plot(table = table_meta,
                  metadata = metadata_meta,
                  taxonomy_db = "Kraken2",
                  level = "genus",
                  top_n = 20,
                  x_col = "Polygon",
                  x_axis_title = "Polygons",
                  add_remained = TRUE,
                  label = "Genus",
                  save_table = FALSE)
```

### 🖥️ Abundance heatmap
Kraken2/Bracken taxa aren't ASVs, so `feature_prefix = "Taxon"` is used here instead of the default `"ASV"` row-label prefix.

```{r, warning=FALSE, fig.width=8.5, fig.height=5.5, out.width="85%"}
abundance_heatmap_plot(table = table_meta,
                  metadata = metadata_meta,
                  top_n = 20,
                  show_column_names = FALSE,
                  condition1 = "Polygon",
                  feature_prefix = "Taxon")
```

## 📊 Alpha diversity
### 📈 Correlation among Hill numbers
```{r, warning=FALSE, fig.width=12, fig.height=4.5}
alpha_hill_corr_plot(table = table_meta)
```

### 📊 Alpha diversity visualization
`fill_col` is the same as `x_col` here, so the legend would just repeat the x-axis tick labels - `show_legend = FALSE` drops it.

```{r, warning=FALSE, fig.width=9, fig.height=5}
alpha_hill_plot(table = table_meta,
                metadata = metadata_meta,
                x_col = "Polygon",
                fill_col = "Polygon",
                facet_by = "Site",
                facet_orientation = "horizontal",
                legend_title = "",
                stat = "kruskal.test",
                show_legend = FALSE,
                save_table = FALSE)
```

Richness (*q*=0) is nearly saturated and flat across blocks here — this genus-level Bracken table simply doesn't have much room left to vary at *q*=0 — while *q*=1 and *q*=2 (which weight by relative abundance) do pick up differences between blocks.

### Alpha diversity along a continuous gradient

```{r, warning=FALSE, fig.height=4, fig.width=10}
alpha_decay_plot(table = table_meta,
                 metadata = metadata_meta,
                 cont_var = "pH",
                 x_axis_title = "Soil pH")
```

Diversity at *q*=1 and *q*=2 increases significantly with soil pH; *q*=0 doesn't, for the same ceiling-effect reason as above.

## Beta diversity

### Ordination of community composition

We can use metrics that consider only presence/absence like **jaccard**:

```{r, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
beta_ord_plot(table = table_meta,
              metadata = metadata_meta,
              distance = "jaccard",
              ordination = "PCoA",
              group_col = "Polygon")
```

### Statistical tests for beta diversity

`method = "jaccard"` matches with the one used for the ordinatrion above, so the ordination (via the `vegan` package [@oksanen2024vegan]) and the PERMANOVA significance test [@anderson2001permanova] are evaluating the same notion of dissimilarity.

```{r, fig.width=8, fig.height=4}
beta_test_table(table = table_meta,
                metadata = metadata_meta,
                formula_str = "Polygon*Site",
                method = "jaccard",
                test = "permanova",
                permutations = 999)
```


### Distance-decay of community similarity

`beta_decay_plot()` tests whether communities that are geographically closer are also more similar in composition, via a Mantel test [@mantel1967test]:

```{r, warning=FALSE, fig.width=5.5, fig.height=4.2, out.width="70%"}
beta_decay_plot(
  table    = table_meta,
  metadata = metadata_meta,
  lat_col  = "Latitude",
  lon_col  = "Longitude",
  distance = "jaccard"
)
```

Communities that are geographically closer are significantly more similar (Mantel *r* ≈ 0.2, *p* = 0.002. the exact value jitters slightly between runs since the Mantel test is permutation-based), a real distance-decay signal across these forest soils. `jaccard` is used to be consistent with the ordination above. 

## Differential abundant analysis

For the heatmap (ALDEx2 [@fernandes2014aldex2]) and ANCOMBC2 [@lin2020ancombc] versions we need a 2-group subset of the 6-level `Polygon`, in this part we will compare Polygon 2 and 5:

```{r}
metadata_meta_compar <- metadata_meta %>%
  filter(Polygon == "Pol2" | Polygon == "Pol5")

table_meta_compar <- table_meta[, c(match(metadata_meta_compar$SAMPLEID, colnames(table_meta)), ncol(table_meta))]
```

```{r, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, out.width="50%"}
aldex_heatmap_plot(table = table_meta_compar,
                   metadata = metadata_meta_compar,
                   col_cond = "Polygon")
```

```{r, fig.width=8, fig.height=6, message=FALSE, warning=FALSE, out.width="50%"}
ancombc_plot(
  table        = table_meta_compar,
  metadata     = metadata_meta_compar,
  col_cond     = "Polygon",
  tax_level    = "Genus",
  prv_cut      = 0.05,
  p_adj_method = "BH"
)
```

### Identifying important taxa

`random_forest_lollipop_plot()` uses a Random Forest model [@breiman2001randomforest] to find the taxa that best predict the polygon:

```{r, fig.width=7, fig.height=6, out.width="60%"}
random_forest_lollipop_plot(table = table_meta, metadata = metadata_meta, top_n = 10, variable_to_predict = "Polygon")
```

```{r, fig.width=8, fig.height=9, out.width="60%"}
ratios_bubble_plot(table = table_meta_compar,
           metadata = metadata_meta_compar,
           condition_col = "Polygon",
           top_n = 20,
           condition_A = "Pol2",
           condition_B = "Pol5")
```

## Environmental analyses

The metadata already carries the soil physicochemical variables measured for these samples, so the environmental table is built the same way as in the metabarcoding vignette:

```{r}
env_table_meta <- metadata_meta %>%
  dplyr::select(SAMPLEID, pH:ARENA) %>%
  remove_rownames() %>%
  column_to_rownames(var = "SAMPLEID")
```

### Correlation between environmental variables and taxonomic abundance

```{r, fig.width=9, fig.height=3.2}
corr_env_abund_plot(table = table_meta,
                    env_table = env_table_meta,
                    metadata = metadata_meta,
                    taxonomy_db = "Kraken2",
                    level = "phylum",
                    save_table = FALSE)
```

### Constrained ordination (CCA)

```{r}
env_table_meta <- env_table_meta[
  match(metadata_meta$SAMPLEID, rownames(env_table_meta)), , drop = FALSE]
```

```{r, fig.width=8, fig.height=6}
cca_rda_biplot(table = table_meta,
               env_data = env_table_meta,
               metadata = metadata_meta,
               env_vars = c("pH", "MO", "N", "P", "K"),
               analysis = "CCA",
               show_all_env_vectors = TRUE,
               group_col = "Polygon",
               scale_arrows = 3)
```


## References

<div id="refs"></div>

## Session info

```{r}
sessionInfo()
```
