---
title: "Metabarcoding"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Metabarcoding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Sensible package-wide defaults for figure size: most plots here are single
# panels or small facet grids, so 7x5in reads well without oversizing text
# relative to the panel (this package's theme uses fixed point sizes, so a
# too-large fig.width/fig.height makes labels look tiny once rendered).
# out.width keeps the display responsive to the page instead of overflowing
# it, independent of each chunk's fig.width/fig.height (which still control
# the plot's internal proportions/aspect ratio).
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

For this example, we will use *metabarcoding* data, specifically 16S for bacteria and 18S for fungi.

The data is taken from:

For 16S: [Hereira-Pacheco, S.E., Navarro-Noya, Y.E. & Dendooven, L. The root endophytic bacterial community of Ricinus communis L. resembles the seeds community more than the rhizosphere bacteria independent of soil water content. Sci Rep 11, 2173 (2021). https://doi.org/10.1038/s41598-021-81551-7](https://doi.org/10.1038/s41598-021-81551-7)

For 18S: [Hereira-Pacheco, S. E., Estrada-Torres, A., Dendooven, L., & Navarro-Noya, Y. E. (2023). Shifts in root-associated fungal communities under drought conditions in Ricinus communis. Fungal Ecology, 63, 101225.](https://www.sciencedirect.com/science/article/abs/pii/S1754504823000028)

First, let's load the `MicroBioMeta` package:

```{r "start", message=FALSE}
library("MicroBioMeta")
```

Let's also load the other packages we'll need:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(qiime2R)
```

[`tidyverse`](https://tidyverse.org/) is used for all the data wrangling and visualizations and [`qiime2R`](https://github.com/jbisanz/qiime2R) for calling `.qza` objects as output from [`QIIME2`](https://qiime2.org/).

Next, we load the data and reformat it. This dataset was processed in [`QIIME2`](https://qiime2.org/), so `read_qza()` is used to import the `.qza` artifacts into R. Alternatively, you could export the same data from `QIIME2` as plain `.txt` files and read them directly.

First, let's load all the raw data:

```{r}

bacteria_table <- qiime2R::read_qza("articles/data/table_bacteria.qza")$data %>% as.data.frame()

fungi_table <- qiime2R::read_qza("articles/data/table_fungi.qza")$data %>%
  as.data.frame()

# The sample-ID column, Treatment labels (TC/TD/TED), and Type_of_soil
# spelling (e.g. Rhizosphere, Bulk soil) are already clean in these files, so
# no renaming or recoding is needed here - just read and, for bacteria,
# keep only the month-2 timepoint.
bacteria_metadata <- read.delim("articles/data/metadata_bacterias.txt", check.names = FALSE) %>%
  filter(Month == "2")

fungi_metadata <- read.delim("articles/data/metadata_fungis.txt", check.names = FALSE)


taxonomy_bacteria <- read_qza("articles/data/taxonomy_bacteria.qza")$data %>%
  as.data.frame() %>% rename(taxonomy = Taxon) %>% 
  dplyr::select(-Confidence) %>% 
  column_to_rownames(var = "Feature.ID")

taxonomy_fungi <- read_qza("articles/data/taxonomy_fungi.qza")$data %>% as.data.frame() %>%
  rename(taxonomy = Taxon) %>% 
  dplyr::select(-Consensus) %>% 
  column_to_rownames(var = "Feature.ID")

```

Then, Let's filter and manage data. It is a good practice to use table and metadata that contains de same information and samples and in the same order.

For bacteria:

```{r}
samples_bac <- bacteria_metadata$SAMPLEID[
  bacteria_metadata$SAMPLEID %in% colnames(bacteria_table)]

table_bacteria <- bacteria_table[, samples_bac]

metadata_bacteria <- bacteria_metadata[
  bacteria_metadata$SAMPLEID %in% samples_bac, ]
```

and for fungi:

```{r}
samples_fun <- fungi_metadata$SAMPLEID[
  fungi_metadata$SAMPLEID %in% colnames(fungi_table)
]

table_fungi <- fungi_table[, samples_fun]

metadata_fungi <- fungi_metadata[
  fungi_metadata$SAMPLEID %in% samples_fun,
]
```

## Pre-processing data

To be used as input for most `MicroBioMeta` functions, the data should be in `biom` format, with a **taxonomy** column at the end.

If your data is not already in that form, `merge_feature_taxonomy()` combines an abundance table and a taxonomy table into this format. Keeping taxonomy as a column of the same table means most functions can look up taxonomic identity without requiring a separate object.

```{r}

table_bac <- merge_feature_taxonomy(table = table_bacteria,
                                    taxonomy = taxonomy_bacteria)

table_fung <- merge_feature_taxonomy(table = table_fungi, 
                                     taxonomy = taxonomy_fungi)

```

## Composition exploration

`MicroBioMeta` has several functions to explore composition, taxonomic identity, and abundance. Let's explore each of them.

### Abundance barplots

Abundance barplots are useful to identify patterns in community composition based on taxonomic identity.\
This function generates stacked barplots of relative abundances using an abundance table and a metadata file as input.

The `taxonomy_db` parameter specifies the taxonomic reference database used for taxonomic assignment (e.g., SILVA, GTDB, UNITE).\
The `x_col` parameter indicates the column in the metadata used for grouping samples along the x-axis, while `facet_col` defines the metadata column used to split the plot into facets.

The `level` parameters allow us to choose the taxonomic level to collapse.

The `top_n_groups` controls the number of most abundant taxa displayed in the barplot.

If `add_remain = TRUE`, the relative abundances are completed to 100% by adding a gray category representing the remaining taxa not explicitly displayed; if `FALSE`, only the selected taxa are shown.\
The `label` parameter defines the legend title displayed in the plot.

The x-axis tick labels can be rotated with `x_label_angle` (default `0`, horizontal) — useful when sample or group names are long enough to overlap, as in the examples below. `strip_text_bold` controls whether facet strip titles are bold (default `FALSE`), and `aspect_ratio` fixes the height/width ratio of the panel when you need consistent proportions across figures.

Finally, when `save_table = TRUE`, the function also returns a table containing the relative abundances (percentages) used to generate the plot.

```{r, fig.width=9, fig.height=6}
abundance_bar_plot(table = table_bac,
                  metadata = metadata_bacteria,
                  taxonomy_db = "silva",
                  level = "phylum",
                  top_n = 15,
                  x_col = "Type_of_soil",
                  facet_col = "Treatment",
                  add_remained = TRUE,
                  label =   "Phylum",
                  x_label_angle = 45,
                  save_table = FALSE)

```

```{r, fig.width=9, fig.height=6}
abundance_bar_plot(table = table_fung,
                  metadata = metadata_fungi,
                  x_col = "Treatment",
                  facet_col = "Type_of_soil",
                  add_remained = TRUE,
                  label =   "Genera",
                  x_label_angle = 45,
                  save_table = FALSE)
```

### Abundance heatmaps

The input `table` must be a data frame containing taxonomic features (rows) and samples (columns).\
The `metadata` argument provides sample-level information used to annotate the heatmap.

The parameters `condition1`, `condition2`, and `condition3` define up to three metadata variables to be displayed as horizontal annotations above the heatmap.\
Custom color palettes for each annotation can be specified using `colors_condition1`, `colors_condition2`, and `colors_condition3`, respectively.\
Legend titles for these annotations can be customized with `name_legend_condition1`, `name_legend_condition2`, and `name_legend_condition3`.

The `top_n` argument controls the number of most abundant features included in the heatmap.\
If `cluster = TRUE`, rows are hierarchically clustered; if `FALSE`, features are ordered by decreasing relative abundance.\
The `show_column_names` parameter determines whether sample names are displayed in the heatmap.

```{r, warning=FALSE, fig.width=8, fig.height=9}
abundance_heatmap_plot(table = table_bac,
                  metadata = metadata_bacteria,
                  top_n = 30,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil",
                  condition2 = "Treatment")
```

```{r, warning=FALSE, fig.width=8, fig.height=9}
abundance_heatmap_plot(table = table_fung,
                  metadata = metadata_fungi,
                  top_n = 30,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil",
                  condition2 = "Treatment")
```

### Sankey diagram of taxonomic flow

`abundance_sankey_plot()` visualizes how relative abundance flows across taxonomic ranks — from Kingdom/Domain down to Species — as an interactive Sankey diagram. It is especially useful for getting a quick overview of which lineages dominate the community without picking a single taxonomic level in advance, the way `abundance_bar_plot()` requires.

The `taxRanks` argument specifies which ranks to include as nodes, `maxn` caps the number of taxa kept per rank (to avoid an unreadable diagram), and `taxonomy_db` tells the function how the taxonomy strings are formatted (e.g. `"silva"`, `"gg"`, `"kraken2"`) so prefixes like `p__`/`g__` are parsed correctly. The diagram is written to an HTML file (`output_file`) that can be opened in a browser or embedded in a report.

This function depends on the [`sankeyD3`](https://github.com/fbreitwieser/sankeyD3) package, which is only available on GitHub:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("fbreitwieser/sankeyD3")
```

```{r, eval=FALSE}
abundance_sankey_plot(
  table       = table_bac,
  output_file = "sankey_bacteria.html",
  maxn        = 15,
  taxRanks    = c("P", "C", "O", "F", "G"),
  taxonomy_db = "silva"
)
```

## Alpha diversity

Alpha diversity describes the diversity within individual samples. In `MicroBioMeta`, alpha diversity can be explored using several metrics, including traditional indices such as richness, Shannon, and Simpson. In addition, the package implements functions based on the **Hill numbers framework**, which provides a unified way to represent these indices as members of the same family of diversity measures. Hill numbers are parameterized by the order *q*, which controls the sensitivity of the metric to species abundances. For example, *q*=0 corresponds to species richness, *q*=1 to the exponential of Shannon entropy, and *q*=2 to the inverse Simpson index. This formulation facilitates consistent comparison among diversity measures and allows researchers to evaluate how diversity patterns change depending on the weight given to rare versus abundant taxa.

### Correlation among Hill numbers

The `alpha_hill_corr_plot()` function calculates Hill numbers for different orders (typically, 𝑞=0,1,2) and visualizes the correlations among them using a correlation plot.

This plot helps evaluate how strongly diversity metrics of different orders are related to the depth sequencing.

The table argument must be a data frame containing features (rows) and samples (columns).

For the bacteria, we can observed that at all orders of *q* there is a strong correlation with sequencing depth.

```{r, warning=FALSE, fig.width=12, fig.height=4.5}
alpha_hill_corr_plot(table = table_bac)

```

But, for fungi data this correlation is just strong and significative at *q*=0, it means for richness but not to the others *q* orders.

```{r, warning=FALSE, fig.width=12, fig.height=4.5}
alpha_hill_corr_plot(table = table_fung)
```

### Alpha diversity visualization

The `alpha_hill_plot()` function calculates Hill numbers and visualizes alpha diversity patterns across experimental conditions using boxplots.

The `metadata` argument provides sample-level information used to group and visualize the samples.

The `x_col` parameter specifies the metadata column used on the x-axis, while `fill_col` controls the grouping used to color the boxplots.

The `facet_by` argument allows splitting the plot according to a metadata variable, and `facet_orientation` controls whether facets are arranged horizontally or vertically.

If `save_table = TRUE`, the function also returns a table containing the alpha diversity values calculated for each sample.

```{r, warning=FALSE, fig.width=8, fig.height=6}
alpha_hill_plot(table = table_bac,
                metadata = metadata_bacteria,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "vertical",
                save_table = FALSE)
```

```{r, warning=FALSE, fig.height=6, fig.width=9}
alpha_hill_plot(table = table_fung,
                metadata = metadata_fungi,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                save_table = FALSE)
```

Also, we can visualize other metrics considering classic alpha diversity indexes. Also, using the parameter stat it is possible to set an statistical comparison indication the method to use:

```{r, warning=FALSE, fig.height=6, fig.width=9}
alpha_diversity_plot(table = table_fung,
                metadata = metadata_fungi,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal", stat = "anova",
                save_table = FALSE)
```

When a statistical comparison (`stat`) is requested, each panel is automatically tagged (A, B, C, ...) so it can be referenced in a figure caption. `panel_label_case` switches between upper- and lowercase tags, `panel_label_bold` toggles bold tags, and `panel_labels` lets you supply custom tags (e.g. `c("(a)", "(b)", "(c)")`) instead of the auto-generated ones. `free_y` lets each panel use its own y-axis scale instead of a shared one — useful when comparing Hill orders (*q*=0, 1, 2) that differ by an order of magnitude, as below:

```{r, warning=FALSE, fig.height=6, fig.width=9}
alpha_diversity_plot(table = table_fung,
                metadata = metadata_fungi,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                stat = "anova",
                free_y = TRUE,
                x_label_angle = 45,
                panel_label_bold = FALSE,
                save_table = FALSE)
```

### Alpha diversity along a continuous gradient

Instead of comparing diversity across discrete groups, `alpha_decay_plot()` regresses Hill numbers against a **continuous** metadata variable (e.g. pH, elevation, distance to a reference point). Each Hill order (*q*=0, 1, 2) gets its own panel with an OLS regression line and an annotation reporting the correlation coefficient (Spearman's rho by default), its p-value, and — when `show_lm_stats = TRUE` — the linear model's R² and slope.

The `cont_var` parameter names the continuous metadata column to place on the x-axis, and the optional `group_col` fits a separate regression line per group instead of a single global one.

Here we use soil `pH` as the gradient:

```{r, warning=FALSE, fig.height=4, fig.width=10}
alpha_decay_plot(table = table_bac,
                 metadata = metadata_bacteria,
                 cont_var = "pH",
                 group_col = "Type_of_soil",
                 x_axis_title = "Soil pH")
```

### Shared taxa between sample groups

Another way to explore alpha diversity patterns is by examining **which taxa are shared or unique among groups of samples**. The function `venn_plot()` generates Venn diagrams that summarize the overlap of taxa between groups defined in the metadata.

The function can optionally filter taxa using a **minimum prevalence threshold**, allowing the visualization to focus only on taxa consistently detected within each group.

Next, we generate a Venn diagram grouping samples according to the soil compartment.

```{r, fig.width=7, fig.height=6}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  min_prevalence = 0 )
```

This plot shows the number of taxa that are **unique to each soil compartment** and those **shared among groups**.

#### Collapsing the table by taxonomic level

Microbial community tables are often generated at the **ASV or OTU level**, which can contain hundreds or thousands of features. For interpretation and visualization, it is often useful to aggregate these features at a higher **taxonomic rank**, such as genus or family.

The function `collapse_table()` performs this aggregation by grouping features that share the same taxonomy at a specified level and summing their abundances across samples. Taxa that do not reach the selected taxonomic resolution are retained without modification.

The function returns a **list with two elements**:

-   **`collapsed_table`**: a wide-format abundance table where rows correspond to taxa collapsed at the selected level and columns correspond to samples.

-   **`long_format`**: the same information in long format, which is useful for downstream plotting or statistical analyses.

-   For example, the following code collapses the fungal abundance table to the **genus level**.

```{r, fig.width=7, fig.height=6}

table_genus <- collapse_table(
  table = table_fung,
  metadata = metadata_fungi,
  level = "genus"
)

venn_plot(
  table = table_genus$collapsed_table,
  metadata = metadata_fungi,
  merge_by = "Type_of_soil"
)
```

By default, `collapse_table()` returns the **raw counts** summed at the selected taxonomic level. However, microbial community data are often easier to interpret when expressed as **relative abundance**, since sequencing depth can vary across samples.

Setting `rel_abun = TRUE` converts the counts of each sample into **percentages**, where the total abundance per sample sums to 100%.

#### Filtering taxa by prevalence

Sometimes it is useful to visualize only taxa that occur frequently within each group. This can be done using the `min_prevalence` parameter.

For example, the following diagram includes only taxa present in **at least 20% of the samples within each group**.

```{r, fig.width=7, fig.height=6}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  min_prevalence = 0.2 )
```

#### Customizing group colors

The function also allows custom colors for the groups.

```{r, fig.width=7, fig.height=6}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  min_prevalence = 0,   
                  group_colors = c("#1B9E77", "#D95F02", "#7570B3") )
```

#### Alternative plotting method

By default the function uses the package **ggvenn**, but it can also generate the diagram using **ggVennDiagram**.

```{r}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  method = "ggvenndiagram" )
```

## Beta diversity

Beta diversity describes the differences in community composition among samples. These analyses help determine whether microbial communities vary across environmental conditions, treatments, or experimental groups.

`MicroBioMeta` provides several tools to explore beta diversity, including ordination methods, statistical tests for community differences, and partitioning approaches that separate shared and unique components of diversity. These functions allow both visual exploration of compositional patterns and formal hypothesis testing.

### Ordination of community composition

The `beta_ord_plot()` function calculates pairwise dissimilarities among samples and visualizes the results using an ordination method. By default, the function uses common ecological distance metrics (e.g., Bray–Curtis or Euclidean) and dimensionality reduction techniques such as **PCoA**, **PCA**, or **NMDS**. Accounting for compositional data in the examples `aitchinson` and `compositional` distances are used. Compositional refers to the *clr* (log centered ratio) transformation made by `ALDEx2` package.

The `group_col` parameter specifies the metadata column used to color the samples, while `shape_col` allows an additional metadata variable to be displayed using different point shapes.

The `distance` argument defines the dissimilarity metric used to calculate beta diversity, and `ordination` specifies the ordination method applied to visualize the distances among samples.

```{r, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
beta_ord_plot(table = table_bac,
              metadata = metadata_bacteria,
               distance = "compositional",
              ordination = "PCA",
              group_col = "Type_of_soil",
              shape_col = "Treatment")
```

```{r, fig.width=8, fig.height=6}
beta_ord_plot(table = table_fung,
              metadata = metadata_fungi,
              group_col = "Type_of_soil",
              shape_col = "Treatment",
              distance = "aitchison",
              ordination = "NMDS")

```

### Statistical tests for beta diversity

The `beta_test_table()` function performs statistical tests to evaluate whether community composition differs among experimental groups.

The `formula_str` argument specifies the experimental design using a formula syntax similar to linear models in R. The `method` parameter defines the distance metric used to calculate dissimilarities, and the `test` argument specifies the statistical method (e.g., **PERMANOVA**).

The `permutations` parameter determines the number of permutations used to estimate statistical significance.

```{r, fig.width=8, fig.height=4}
beta_test_table(table = table_bac,
                metadata= metadata_bacteria,
                formula_str = "Type_of_soil*Treatment",
                method = "euclidean",
                test = "permanova",
                permutations = 999)
```

```{r, fig.width=8, fig.height=4}
beta_test_table(table = table_fung,
                metadata= metadata_fungi,
                formula_str = "Type_of_soil*Treatment",
                method = "euclidean",
                test = "permanova",
                permutations = 999)
```

### Beta diversity partitioning

Beta diversity can also be decomposed into different components that describe how communities differ. The `beta_partition_ord_plot()` function partitions beta diversity into shared and unique fractions based on selected dissimilarity indices.

The `index` argument specifies the dissimilarity family used for the partition (e.g., Jaccard or Sørensen). The `group_col` and `shape_col` parameters define how samples are visualized according to metadata variables.

```{r, fig.width=14, fig.height=5.5}
beta_partition_ord_plot(
  table = table_bac,
  metadata = metadata_bacteria,
  index = "jaccard",
  group_col = "Type_of_soil",
  shape_col = "Treatment", 
  save_table = FALSE)
```

```{r, fig.width=14, fig.height=5.5}
beta_partition_ord_plot(
  table = table_fung,
  metadata = metadata_fungi,
  index = "jaccard",
  group_col = "Type_of_soil",
  shape_col = "Treatment", 
  save_table = FALSE)
```

### Pairwise beta diversity visualization

The `beta_dissimilarity_plot()` function summarizes pairwise beta diversity values using boxplots, allowing comparison of diversity patterns among groups of samples.

The `condition1_col` and `condition2_col` parameters specify metadata variables used to define the groups being compared. The `partition` argument determines which component of beta diversity is displayed (e.g., shared or unique), while `family` defines the dissimilarity index used.

By default `beta_dissimilarity_plot()` keeps every pairwise combination of `condition1_col`, including a group compared against itself (e.g. `"Roots_vs_Roots"`), which isn't usually meaningful. The `comparison_condition1` argument restricts the plot to specific `"value1_vs_value2"` labels (the two values sorted alphabetically and joined with `"_vs_"`). Here we build that list programmatically to keep only the comparisons of every other compartment **against Roots**, dropping self-comparisons:

```{r}
groups_bac <- setdiff(unique(as.character(metadata_bacteria$Type_of_soil)), "Roots")
vs_roots_bac <- paste0(pmin(groups_bac, "Roots"), "_vs_", pmax(groups_bac, "Roots"))
vs_roots_bac
```

As with the alpha diversity plots, `x_label_angle` rotates the x-axis tick labels (default `0`), `strip_text_bold` toggles bold facet strip titles (default `FALSE`), and `aspect_ratio` fixes the panel's height/width ratio. As with the alpha plots, `stat` (e.g. `"wilcox.test"`, `"kruskal.test"`, `"anova"`) adds a statistical comparison across the boxplots, passed through to `ggpubr::stat_compare_means()`.

If `save_table = TRUE`, the function also returns a table containing the beta diversity values used to generate the plot.

```{r, fig.width=9, fig.height=6}
beta_dissimilarity_plot(table = table_bac,
                       metadata = metadata_bacteria,
                       comparison_condition1 = vs_roots_bac,
                       condition1_col = "Type_of_soil",
                       condition2_col = "Treatment",
                       x_axis_title = "Samples",
                       x_label_angle = 45,
                       stat = "kruskal.test",
                       partition = "shared", family = "jaccard",
                       save_table = FALSE)

```

```{r}
groups_fung <- setdiff(unique(as.character(metadata_fungi$Type_of_soil)), "Roots")
vs_roots_fung <- paste0(pmin(groups_fung, "Roots"), "_vs_", pmax(groups_fung, "Roots"))
```

```{r, fig.width=9, fig.height=6}
beta_dissimilarity_plot(table = table_fung,
                       metadata = metadata_fungi,
                       comparison_condition1 = vs_roots_fung,
                       condition1_col = "Type_of_soil",
                       condition2_col = "Treatment",
                       x_axis_title = "Samples",
                       x_label_angle = 45,
                       stat = "kruskal.test",
                       partition = "shared",
                       family = "jaccard",
                       save_table = FALSE)

```

### Pairwise turnover: between-group vs. within-group

`beta_turnover_plot()` answers a different question than `beta_dissimilarity_plot()`: for **one reference group** (e.g. Rhizosphere), is its ASV turnover **against another group** (e.g. Roots — a *between-group* comparison) greater than its turnover **against itself** (Rhizosphere-vs-Rhizosphere pairs — a *within-group* baseline)? If between-group turnover is significantly higher, that's evidence the two compartments really do host distinct communities, beyond the variability already present among replicates of the same compartment. It reports the ASV turnover component of Hill-number-based beta diversity (*q*=0, 1, 2), one facet row per Hill order.

`comparison_condition1` selects which `"value1_vs_value2"` combinations (built internally from `condition1.x`/`condition1.y`) to keep — unlike `beta_dissimilarity_plot()`, these are **not** sorted alphabetically, so `"Rhizosphere_vs_Roots"` (between-group) and `"Rhizosphere_vs_Rhizosphere"` (within-group) both need to be listed explicitly, with Rhizosphere always first. `comparison_condition2` keeps only same-treatment pairs, so turnover isn't confounded by comparing samples from different treatments.

`color_facets_x` colors the top facet strip (`condition1.x` is constant here, "Rhizosphere", since it's the reference group in both comparisons) and `color_axis_x` colors the two boxes (the between- and within-group values) on the x-axis (`condition1.y`). Tick labels are hidden by default since the groups are already identified in the legend — set `show_x_labels = TRUE` to bring them back. `stat` (e.g. `"wilcox.test"`) adds a statistical comparison between the two boxes.

```{r, warning=FALSE, fig.width=6, fig.height=9}
beta_turnover_plot(
  table                 = table_fungi,
  metadata              = metadata_fungi,
  comparison_condition1 = c("Rhizosphere_vs_Roots", "Rhizosphere_vs_Rhizosphere"),
  comparison_condition2 = c("TC_vs_TC", "TD_vs_TD", "TED_vs_TED"),
  condition1.x          = "Type_of_soil.x",
  condition1.y          = "Type_of_soil.y",
  condition2.x          = "Treatment.x",
  condition2.y          = "Treatment.y",
  color_facets_x        = "#5D478B",
  color_axis_x          = c("Roots" = "#56B4E9", "Rhizosphere" = "#E69F00"),
  stat                  = "wilcox.test"
)$plot
```

### Distance-decay of community similarity

`beta_decay_plot()` tests whether communities that are geographically closer are also more similar in composition — the classic **distance-decay** relationship. It computes pairwise community dissimilarity (`distance`, e.g. `"jaccard"`, `"bray"`, `"horn"`) and pairwise geographic distances from sample coordinates, runs a Mantel test between them, and plots similarity (1 − dissimilarity) against distance with a fitted regression line.

`lat_col`/`lon_col` name the latitude/longitude columns in `metadata` (decimal degrees), and the optional `group_col` runs a separate Mantel test and regression line per group instead of one global test.

Our example metadata doesn't include real coordinates, so here we build small synthetic latitude/longitude values per sampling site (`Loc`) purely to demonstrate the function — substitute your own site coordinates for a real analysis.

```{r, warning=FALSE, fig.width=8, fig.height=6}
loc_coords <- data.frame(
  Loc = 0:8,
  lat = 19.0 + seq(0, 0.8, length.out = 9),
  lon = -99.0 + seq(0, 0.8, length.out = 9)
)
metadata_bacteria$lat <- loc_coords$lat[match(metadata_bacteria$Loc, loc_coords$Loc)]
metadata_bacteria$lon <- loc_coords$lon[match(metadata_bacteria$Loc, loc_coords$Loc)]

# "Uncultivated" has too few samples for a per-group Mantel test to return
# anything (it comes back as NA and just clutters the plot), so it's
# excluded from this particular example - group_col works fine with groups
# this small as long as each one actually has enough pairs to test.
metadata_bacteria_decay <- metadata_bacteria %>%
  filter(Type_of_soil != "Uncultivated")

beta_decay_plot(
  table     = table_bac,
  metadata  = metadata_bacteria_decay,
  lat_col   = "lat",
  lon_col   = "lon",
  distance  = "jaccard",
  group_col = "Type_of_soil"
)
```

## Differential abundant analysis

`MicroBioMeta` provides several functions to explore **differentially abundant taxa between conditions**. These analyses help identify microbial features that vary significantly across environments or experimental groups.

Two functions rely on a **compositional data analysis framework implemented in the package** [ALDEx2]{.underline}. This approach accounts for the compositional nature of sequencing data by using Monte Carlo Dirichlet instances and centered log-ratio transformations, making it more appropriate for high-throughput sequencing datasets than traditional count-based methods.

The functions:

-   `aldex_volcano_plot()` visualizes differential abundance results using a **volcano plot**.

-   `aldex_heatmap_plot()` highlights significant taxa in a **heatmap representation**.

-   `ancombc_plot()` runs [ANCOMBC2](https://bioconductor.org/packages/ANCOMBC/), an alternative differential abundance method that explicitly models compositional bias and structural zeros, and returns a bar plot (2 groups), a heatmap (3+ groups), or a bar plot of the effect size per unit increase (continuous covariate).

-   Additionally, two complementary approaches help identify **important taxa that discriminate between groups**:

-   `ratios_bubble_plot()` highlights taxa based on **log-ratio differences between conditions**.

-   `random_forest_lollipop_plot()` uses a **Random Forest model** to identify features that best predict a given variable, visualized as a **lollipop plot of feature importance**.

-   Since ALDEx2 performs **pairwise comparisons**, we first subset the data to include only two experimental conditions.

### Example: Bacterial data

First, we filter the metadata to retain only two soil compartments.

```{r}
metadata_bacteria_compar <-  metadata_bacteria %>% 
  filter(Type_of_soil == "Roots" | Type_of_soil =="Rhizosphere")

table_bacteria_compar <- table_bacteria[match(metadata_bacteria_compar$SAMPLEID, colnames(table_bacteria))]

table_bac_compar <- merge_feature_taxonomy(table_bacteria_compar, taxonomy_bacteria)

```

We can visualize differential abundance using a **volcano plot**. Taxa near the significance thresholds are labeled with their name; `label_size` controls the font size of these labels, and `filter_uncultured = TRUE` drops "uncultured"/"unculture" taxa from the labels (they are still plotted as points, just not labeled) — useful when such generic labels would otherwise crowd out more informative ones.

```{r, fig.width=7, fig.height=5.5, message=FALSE, warning=FALSE}
aldex_volcano_plot(table = table_bac_compar,
                   metadata = metadata_bacteria_compar,
                   col_cond = "Type_of_soil",
                   type = "volcano",
                   label_size = 3,
                   filter_uncultured = TRUE )

```

### ANCOMBC2 as an alternative method

`ancombc_plot()` provides an alternative to the ALDEx2-based functions above. It depends on the Bioconductor packages `ANCOMBC` and `phyloseq`, plus `qiime2R` (GitHub-only):

```{r, eval=FALSE}
# install.packages("BiocManager")
BiocManager::install(c("ANCOMBC", "phyloseq"))
remotes::install_github("jbisanz/qiime2R")
```

For a two-group comparison such as `Type_of_soil` (Roots vs. Rhizosphere), `ancombc_plot()` returns a bar plot of log-fold changes, in the same spirit as the volcano/heatmap plots above but based on ANCOMBC2's own significance testing. `tax_level` agglomerates the ASV table to a taxonomic rank before testing (default `"Genus"`), and `p_adj_method`/`prv_cut` control the multiple-testing correction and prevalence filter.

```{r, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
ancombc_plot(
  table        = table_bac_compar,
  metadata     = metadata_bacteria_compar,
  col_cond     = "Type_of_soil",
  tax_level    = "Genus",
  prv_cut      = 0.05,
  p_adj_method = "BH"
)
```

### Example: Fungal data

We follow the same procedure but visualize the results with a **heatmap**. Unlike the `ggplot2`-based figures above, this heatmap is built with `ComplexHeatmap`, which sizes each row/column in fixed physical units rather than stretching to fill the chunk's `fig.width`/`fig.height` — so its rendered size scales with the **number of significant taxa found**, and will look compact (with blank margin around it) whenever few taxa pass the significance threshold, as is the case here.

```{r}
metadata_fungi_compar <-  metadata_fungi %>%
  filter(Type_of_soil == "Bulk soil" | Type_of_soil =="Rhizosphere")

table_fungi_compar <- table_fung[match(metadata_fungi_compar$SAMPLEID, colnames(table_fung))]

table_fung_compar <- merge_feature_taxonomy(table_fungi_compar, taxonomy_fungi)

```

```{r, fig.width=9, fig.height=4, message=FALSE, warning=FALSE}
aldex_heatmap_plot(table = table_fung_compar,
                   metadata = metadata_fungi_compar,
                   col_cond = "Type_of_soil")

```

### Identifying important taxa

The functions `random_forest_lollipop_plot()` and `ratios_bubble_plot()` can also be used to highlight taxa that contribute most strongly to the observed differences between conditions.

For example, a **Random Forest model** can be used to identify the taxa that best predict the soil compartment:

```{r, fig.width=7, fig.height=6}
random_forest_lollipop_plot(table = table_bac, metadata = metadata_bacteria, top_n = 10, variable_to_predict = "Type_of_soil")
```

Alternatively, taxa with the strongest **log-ratio differences between two conditions** can be visualized using:

```{r, fig.width=8, fig.height=9}
ratios_bubble_plot(table = table_fung_compar,
           metadata = metadata_fungi_compar,
           condition_col = "Type_of_soil",
           top_n = 20,
           condition_A = "Rhizosphere",
           condition_B = "Bulk soil")
```

## Environmental analyses

MicroBioMeta provides two functions to explore the relationship between **environmental variables** and **microbial community composition**:

-   `corr_env_abund_plot()` — computes correlations between taxonomic abundance and environmental variables.

-   `cca_rda_biplot()` — performs constrained ordination using **Canonical Correspondence Analysis (CCA)** or **Redundancy Analysis (RDA)**.

#### Preparing the environmental table

Environmental variables must be organized in a **data frame where rows correspond to samples and columns correspond to environmental variables**.

In this example, environmental variables are extracted from the metadata — here we use the fungal dataset, which has the same set of soil measurements.

```{r}
env_table_fung <- metadata_fungi %>%
  dplyr::select(SAMPLEID, pH:Arbus_per) %>%
  remove_rownames() %>%
  column_to_rownames(var = "SAMPLEID")
```

#### Correlation between environmental variables and taxonomic abundance

The function `corr_env_abund_plot()` calculates correlations between **environmental variables** and **taxonomic relative abundances** at a selected taxonomic level.

Example using **phylum-level abundances**:

```{r, fig.width=9, fig.height=8}
corr_env_abund_plot(table = table_fung,
                    env_table = env_table_fung,
                    metadata = metadata_fungi,
                    level = "phylum",
                    save_table = FALSE)

```

This function produces a **correlation heatmap** showing how environmental variables relate to taxonomic abundances.

### Alternative visualization

Correlations can also be displayed using circle markers instead of tiles:

```{r, fig.width=9, fig.height=8}
corr_env_abund_plot(table = table_fung,
                    env_table = env_table_fung,
                    metadata = metadata_fungi,
                    level = "phylum",
                    save_table = FALSE,
                    geom = "circle")

```

#### Constrained ordination (CCA / RDA)

The function `cca_rda_biplot()` performs a constrained ordination analysis to evaluate how environmental variables explain variation in microbial community composition.

Before running the analysis, ensure that the **sample order in the environmental table matches the metadata**. `env_table_fung` already has `SAMPLEID` as row names (set earlier via `column_to_rownames()`), so we just reorder its rows to match `metadata_fungi`:

```{r}
env_table_fung <- env_table_fung[
  match(metadata_fungi$SAMPLEID, rownames(env_table_fung)), , drop = FALSE]

```

Then run the RDA analysis. The environmental vectors are often short relative to the sample scores (as the arrows themselves only encode *direction*, not the scale shown on the axes) — `scale_arrows` stretches them by a fixed multiplier purely for legibility, without changing their direction or relative lengths:

```{r, fig.width=8, fig.height=6}
cca_rda_biplot(table = table_fung,
               env_data = env_table_fung,
               metadata = metadata_fungi,
               env_vars = c("pH","TN","WHC", "EC", "Clay"),
               analysis = "RDA",
               show_all_env_vectors = TRUE,
               group_col = "Type_of_soil",
               scale_arrows = 3)

```

This analysis produces a **biplot** where:

-   **Points** represent samples

-   **Colors** indicate sample groups defined in the metadata.

-   **Arrows** represent environmental variables and their direction of influence on community structure.

-   Environmental vectors pointing in similar directions indicate **positive associations**, while vectors pointing in opposite directions suggest **negative relationships**
