---
title: "Metabarcoding"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Metabarcoding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

🧬 MicroBioMeta is an R designed to handle metabarcoding data. So, this tutorial walks through an example of its use.

For this example, we will use data sequenced using the 16S barcode for bacteria and 18S barcode for fungi.

The data is taken from:

For 16S 🦠: [Hereira-Pacheco, S.E., Navarro-Noya, Y.E. & Dendooven, L. The root endophytic bacterial community of Ricinus communis L. resembles the seeds community more than the rhizosphere bacteria independent of soil water content. Sci Rep 11, 2173 (2021). https://doi.org/10.1038/s41598-021-81551-7](https://doi.org/10.1038/s41598-021-81551-7)

For 18S 🍄: [Hereira-Pacheco, S. E., Estrada-Torres, A., Dendooven, L., & Navarro-Noya, Y. E. (2023). Shifts in root-associated fungal communities under drought conditions in Ricinus communis. Fungal Ecology, 63, 101225.](https://www.sciencedirect.com/science/article/abs/pii/S1754504823000028)

## 💻 Loading data
First, let's load the `MicroBioMeta` package:

```{r "start", message=FALSE}
library("MicroBioMeta")
```

Let's also load the other packages we'll need:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```

[`tidyverse`](https://tidyverse.org/) is used for all the data wrangling and visualizations.

Next, we load the data and reformat it. This dataset was processed in [`QIIME2`](https://qiime2.org/), so its `.qza` artifacts need to be imported into R. Alternatively, you could export the same data from `QIIME2` as plain `.txt` files and read them directly.

We use in this example: [`qiime2R`](https://github.com/jbisanz/qiime2R) package, whose `read_qza()` this is adapted from (MIT License, © 2018 Jordan Bisanz). For installing:

```{r, eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)){install.packages("devtools")}
devtools::install_github("jbisanz/qiime2R")
```

Let's call it:
```{r}
library(qiime2R)
```

First, let's load all the table and taxonomy data for each data set:

```{r}
bacteria_table <- read_qza("articles/data/table_bacteria.qza")$data %>% as.data.frame()

fungi_table <- read_qza("articles/data/table_fungi.qza")$data %>%
  as.data.frame()

taxonomy_bacteria <- read_qza("articles/data/taxonomy_bacteria.qza")$data %>%
  as.data.frame() %>% rename(taxonomy = Taxon) %>%
  dplyr::select(-Confidence) %>%
  column_to_rownames(var = "Feature.ID")

taxonomy_fungi <- read_qza("articles/data/taxonomy_fungi.qza")$data %>% as.data.frame() %>%
  rename(taxonomy = Taxon) %>%
  dplyr::select(-Consensus) %>%
  column_to_rownames(var = "Feature.ID")

```

⚠️ Let's notice that we renamed the column called Taxon for taxonomy before join this two data:

```{r}
colnames(taxonomy_bacteria)
```


Now, let's call the metadata for data set:

```{r}
bacteria_metadata <- read.delim("articles/data/metadata_bacterias.txt", check.names = FALSE) %>%
  filter(Month == "2")

fungi_metadata <- read.delim("articles/data/metadata_fungis.txt", check.names = FALSE)

```



Then, Let's filter and manage data. It is a good practice to use table and metadata that contains de same information and samples and in the same order. So basically at this point, we are going to filter tables to match with the metadata. 

For bacteria:

```{r}
samples_bac <- bacteria_metadata$SAMPLEID[
  bacteria_metadata$SAMPLEID %in% colnames(bacteria_table)]

table_bacteria <- bacteria_table[, samples_bac]

metadata_bacteria <- bacteria_metadata[
  bacteria_metadata$SAMPLEID %in% samples_bac, ]
```

and for fungi:

```{r}
samples_fun <- fungi_metadata$SAMPLEID[
  fungi_metadata$SAMPLEID %in% colnames(fungi_table)
]

table_fungi <- fungi_table[, samples_fun]

metadata_fungi <- fungi_metadata[
  fungi_metadata$SAMPLEID %in% samples_fun,
]
```

## Pre-processing data

`MicroBioMeta` function needs the data should be in `biom` format, with a **taxonomy** column at the end.

If your data is not already in that form, `merge_feature_taxonomy()` combines an abundance table and a taxonomy table into this format. Keeping taxonomy as a column of the same table means most functions can look up taxonomic identity without requiring a separate object.

```{r}

table_bac <- merge_feature_taxonomy(table = table_bacteria,
                                    taxonomy = taxonomy_bacteria)

table_fung <- merge_feature_taxonomy(table = table_fungi, 
                                     taxonomy = taxonomy_fungi)

```

⚠️ Notice that this warning reports how many taxonomy IDs did not match the feature table and were excluded from the result.


## 📊 Composition exploration
`MicroBioMeta` has several functions to explore composition, taxonomic identity, and abundance. Let's explore each of them.

### 📊 Abundance barplots
Abundance barplots are useful to identify patterns in community composition based on taxonomic identity.\
This function generates stacked barplots of relative abundances using an abundance table and a metadata file as input.

The `taxonomy_db` parameter specifies the taxonomic reference database used for taxonomic assignment (e.g., SILVA, GTDB, UNITE).\
The `x_col` parameter indicates the column in the metadata used for grouping samples along the x-axis, while `facet_col` defines the metadata column used to split the plot into facets.

The `level` parameters allow us to choose the taxonomic level to collapse.

The `top_n_groups` controls the number of most abundant taxa displayed in the barplot.

If `add_remain = TRUE`, the relative abundances are completed to 100% by adding a gray category representing the remaining taxa not explicitly displayed; if `FALSE`, only the selected taxa are shown.\
The `label` parameter defines the legend title displayed in the plot.

The x-axis tick labels can be rotated with `x_label_angle` (default `0`, horizontal), useful when sample or group names are long enough to overlap, as in the examples below. `strip_text_bold` controls whether facet strip titles are bold (default `FALSE`), and `aspect_ratio` fixes the height/width ratio of the panel when you need consistent proportions across figures.

Finally, when `save_table = TRUE`, the function also returns a table containing the relative abundances (percentages) used to generate the plot.

```{r, fig.width=9, fig.height=6}
abundance_bar_plot(table = table_bac,
                  metadata = metadata_bacteria,
                  taxonomy_db = "silva",
                  level = "phylum",
                  top_n = 15,
                  x_col = "Type_of_soil",
                  facet_col = "Treatment",
                  add_remained = TRUE,
                  label =   "Phylum",
                  x_label_angle = 45,
                  save_table = FALSE)

```

```{r, fig.width=9, fig.height=6}
abundance_bar_plot(table = table_fung,
                  metadata = metadata_fungi,
                  x_col = "Treatment",
                  facet_col = "Type_of_soil",
                  add_remained = TRUE,
                  label =   "Genera",
                  x_label_angle = 45,
                  save_table = FALSE)
```

### 🖥️ Abundance heatmaps
Notice that the input files are the same as the previous function and this is consistent with all functions from now on. 

The parameters `condition1`, `condition2`, and `condition3` define the three metadata variables to be displayed as horizontal annotations above the heatmap.\
Custom color palettes for each annotation can be specified using `colors_condition1`, `colors_condition2`, and `colors_condition3`, respectively.\
Legend titles for these annotations can be customized with `name_legend_condition1`, `name_legend_condition2`, and `name_legend_condition3`.

The `top_n` argument controls the number of most abundant features included in the heatmap.\
If `cluster = TRUE`, rows are hierarchically clustered; if `FALSE`, features are ordered by decreasing relative abundance.\
The `show_column_names` parameter determines whether sample names are displayed in the heatmap.

Row labels default to just the feature's rank number (`"1"`, `"2"`...); `feature_prefix` prepends whatever fits the data - `"ASV"` here, since this table's features are QIIME2/DADA2 amplicon sequence variants.

```{r, warning=FALSE, fig.width=11, fig.height=6.5}
abundance_heatmap_plot(table = table_bac,
                  metadata = metadata_bacteria,
                  top_n = 20,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil",
                  condition2 = "Treatment",
                  feature_prefix = "ASV")
```

```{r, warning=FALSE, fig.width=11, fig.height=6.5}
abundance_heatmap_plot(table = table_fung,
                  metadata = metadata_fungi,
                  top_n = 20,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil",
                  condition2 = "Treatment",
                  feature_prefix = "ASV")
```

### 🖥️ Sankey diagram of taxonomic flow
`abundance_sankey_plot()` visualizes how relative abundance flows across taxonomic ranks: from Kingdom to Species as an interactive Sankey diagram. It is useful for getting a quick overview of which lineages dominate the community.

The `taxRanks` argument specifies which ranks to include as nodes, `maxn` caps the number of taxa kept per rank and `taxonomy_db` tells the function how the taxonomy strings are formatted (e.g. `"silva"`, `"gg"`, `"kraken2"`). This function depends on [`networkD3`](https://CRAN.R-project.org/package=networkD3), available on CRAN:

```{r, eval=FALSE}
install.packages("networkD3")
```

⚠️ Unlike the other functions, this one doesn't return a `ggplot` object but a `sankeyNetwork` object  and it has a parameter (`output_file`)  that allows to  **saves an HTML file** with an interactive sankey diafgram.

```{r, warning=FALSE}
sankey_bacteria <- abundance_sankey_plot(
  table       = table_bac,
  output_file = "sankey_bacteria.html",
  maxn        = 15,
  taxRanks    = c("P", "C", "O", "F", "G"),
  taxonomy_db = "silva"
)
sankey_bacteria
```

## 📊 Alpha diversity
Alpha diversity describes the diversity within individual samples. With `MicroBioMeta`, alpha diversity can be explored using several different metrics, like classic metrcis as Chao and Simpson. Also, the package implements functions based on the **Hill numbers framework**, which provides a unified way to represent different units. Hill numbers are parameterized by the order *q*, which regulates the weight given to the abundance. For example, *q*=0 corresponds to species richness (total species), *q*=1 to the exponential of Shannon entropy (frequent species), and *q*=2 to the inverse Simpson index (dominant species). 

### 📈 Correlation among Hill numbers
The `alpha_hill_corr_plot()` function calculates Hill numbers for different orders (typically, 𝑞=0,1,2) and visualizes the correlations among them using a correlation plot and evaluate how strongly diversity metrics of different orders are related to the depth sequencing.

This is the first step before analyze alpha diversity metrics.

For the bacteria 🦠, we can observed that at all orders of *q* there is a strong correlation with sequencing depth.

```{r, warning=FALSE, fig.width=12, fig.height=4.5}
alpha_hill_corr_plot(table = table_bac)

```

But, for fungi 🍄 data this correlation is just strong and significative at *q*=0, it means for richness but not to the others *q* orders.

```{r, warning=FALSE, fig.width=12, fig.height=4.5}
alpha_hill_corr_plot(table = table_fung)
```

### 📊 Alpha diversity visualization
The `alpha_hill_plot()` function calculates Hill numbers and visualizes alpha diversity patterns across experimental conditions using boxplots.

The `x_col` parameter specifies the metadata column used on the x-axis, while `fill_col` controls the grouping used to color the boxplots.

The `facet_by` argument allows splitting the plot according to a metadata variable, and `facet_orientation` controls whether facets are arranged horizontally or vertically.

If `save_table = TRUE`, the function also returns a table containing the alpha diversity values calculated for each sample.

```{r, warning=FALSE, fig.width=8, fig.height=6}
alpha_hill_plot(table = table_bac,
                metadata = metadata_bacteria,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                save_table = FALSE)
```

```{r, warning=FALSE, fig.height=6, fig.width=9}
alpha_hill_plot(table = table_fung,
                metadata = metadata_fungi,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                save_table = FALSE)
```

Also, we can visualize other metrics considering classic alpha diversity indexes:

```{r, warning=FALSE, fig.height=6, fig.width=9}
alpha_diversity_plot(table = table_fung,
                metadata = metadata_fungi,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                stat = "anova",
                free_y = TRUE,
                save_table = FALSE)
```

Each panel is automatically labeled (A, B, C, ...) so it can be referenced. When `stat` is set, a p-value annotation is added to each panel as well (default Wilcoxon for two comparisons and Kruskal-Wallis for three comparisons), we can indicate the method to use with `stat` as the function `stat_compare_means()` of `ggpubr`.
`panel_label_case` switches between upper and lowercase labels, `panel_label_bold` for bold labels and `panel_labels` lets you supply custom labels(e.g. `c("(a)", "(b)", "(c)")`) instead of the auto-generated ones. `free_y` lets each panel use its own y-axis scale instead of a shared one, as below:

```{r, warning=FALSE, fig.height=6, fig.width=9}
alpha_diversity_plot(table = table_fung,
                metadata = metadata_fungi,
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                stat = "anova",
                free_y = TRUE,
                x_label_angle = 45,
                panel_label_bold = FALSE,
                save_table = FALSE)
```

### Alpha diversity along a continuous gradient

 `alpha_decay_plot()` makes a regression Hill numbers against a **continuous** metadata variable (e.g. pH, elevation, distance). Each Hill order (*q*=0, 1, 2) gets its own panel with a regression line and an annotation reporting the correlation coefficient (Spearman's rho by default), its p-value and  when `show_lm_stats = TRUE`, the linear model's R² and slope.

The `cont_var` parameter names the continuous metadata column to place on the x-axis, and the optional `group_col` fits a separate regression line per group instead of a single global one.

Here we use soil `pH` as the gradient, restricted to `Bulk soil` and `Rhizosphere` for a cleaner comparison:

```{r, warning=FALSE, fig.height=4, fig.width=10}
metadata_bacteria_decay <- metadata_bacteria %>%
  filter(Type_of_soil %in% c("Bulk soil", "Rhizosphere"))

alpha_decay_plot(table = table_bac,
                 metadata = metadata_bacteria_decay,
                 cont_var = "pH",
                 group_col = "Type_of_soil",
                 x_axis_title = "Soil pH")
```

### 💻 Shared taxa between sample groups
Another way to explore alpha diversity patterns is by examining **which taxa are shared or unique among groups of samples**. The function `venn_plot()` generates Venn diagrams that summarize the overlap of taxa between groups defined in the metadata.

We generate a Venn diagram grouping samples according to the soil compartment.

```{r, fig.width=7, fig.height=6}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  min_prevalence = 0 )
```



#### 💻 Collapsing the table by taxonomic level
Microbial community tables are often generated at the **ASV or OTU level**, which can contain hundreds or thousands of features. For interpretation and visualization, it is often useful to group these features at a higher **taxonomic level**, such as genus. 

The function `collapse_table()` performs this aggregation by grouping features that are at the same taxonomy level and summing their counts. 

The function returns a **list with two elements**:

-   **`collapsed_table`**: a wide-format abundance table where rows correspond to taxa collapsed at the selected level and columns correspond to samples.

-   **`long_format`**: the same information in long format, which is useful for downstream plotting or statistical analyses.

-   For example, the following code collapses the fungal abundance table to the **genus level** and we make a venn diagram to visualize genera shared.

```{r, fig.width=7, fig.height=6}

table_genus <- collapse_table(
  table = table_fung,
  metadata = metadata_fungi,
  level = "genus"
)

venn_plot(
  table = table_genus$collapsed_table,
  metadata = metadata_fungi,
  merge_by = "Type_of_soil"
)
```

By default, `collapse_table()` returns the **raw counts** summed at the selected taxonomic level. However, microbial community data are needed to visualized or analyzed as **relative abundance**. Setting `rel_abun = TRUE` converts the counts of each sample into **percentages**, where the total abundance per sample sums to 100%.


#### 🖥️ Filtering taxa by prevalence
Sometimes it is useful to visualize only taxa that occur frequently within each group. This can be done using the `min_prevalence` parameter.

For example, the following diagram includes only taxa present in **at least 20% of the samples within each group**. The parameter is default to 0, that means that nor filtering is applied. 

```{r, fig.width=7, fig.height=6}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  min_prevalence = 0.2 )
```

#### 🖥️ Customizing group colors
The function also allows custom colors for the groups.

```{r, fig.width=7, fig.height=6}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  min_prevalence = 0,   
                  group_colors = c("#1B9E77", "#D95F02", "#7570B3") )
```

#### 🖥️ Alternative plotting method
By default the function uses the package **ggvenn**, but it can also generate the diagram using **ggVennDiagram**.

```{r}
venn_plot(table = table_fung,   
                  metadata = metadata_fungi,   
                  merge_by = "Type_of_soil",   
                  method = "ggvenndiagram" )
```

## 💻 Beta diversity
Beta diversity describes the differences in community composition and structure among samples. These analyses help determine whether microbial communities vary across environmental conditions, treatments, or experimental groups and how this clustered or grouped for. 

`MicroBioMeta` has several functions to explore beta diversity, including ordination plots, statistical tests for community differences and partitioning approaches that separate shared and unique components of diversity. These functions allow both visual exploration and formal hypothesis testing.

### 🖥️ Ordination of community composition
The `beta_ord_plot()` function calculates pairwise dissimilarities among samples and visualizes the results using an ordination method. By default, the function uses common ecological distance metrics (e.g., Bray–Curtis or Euclidean) and dimensionality reduction techniques such as **PCoA**, **PCA**, or **NMDS**. Accounting for compositional data in the examples `aitchinson` and `compositional` distances are used. Compositional uses the *clr* (log centered ratio) transformation made by `ALDEx2` package.

The `group_col` parameter specifies the metadata column used to color the samples, while `shape_col` allows an additional metadata variable to be displayed using different point shapes.

The `distance` argument defines the dissimilarity metric used to calculate beta diversity (e.g, **bray, jaccard, aitchsion**, etc), and `ordination` (options: **PCA, PCoA** or **NMDS**) specifies the ordination method applied to visualize the distances among samples.

```{r, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
beta_ord_plot(table = table_bac,
              metadata = metadata_bacteria,
               distance = "compositional",
              ordination = "PCA",
              group_col = "Type_of_soil",
              shape_col = "Treatment")
```

```{r, fig.width=8, fig.height=6}
beta_ord_plot(table = table_fung,
              metadata = metadata_fungi,
              group_col = "Type_of_soil",
              shape_col = "Treatment",
              distance = "compositional",
              ordination = "PCA")

```

### 🖥️ Statistical tests for beta diversity
The `beta_test_table()` function performs statistical tests to evaluate whether community composition differs among experimental groups.

The `formula_str` argument specifies the experimental design using a formula syntax similar to linear models in R or as it is used for `vegan` package. The `method` parameter defines the distance metric used to calculate dissimilarities, and the `test` argument specifies the statistical test (options: **PERMANOVA** or **BETADISPER**).

The `permutations` parameter determines the number of permutations used to estimate statistical significance. The `strata_var` is used as in `vegan` package to block for a parameter. 

⚠️ **NOTE: Each function has the option save_* that export a table or an element obtain in each function**. 

`method` is set to `compositional` for both datasets, matching each one's PCA above, so the significance test evaluates the same notion of dissimilarity shown in the plot.

```{r, fig.width=8, fig.height=4}
beta_test_table(table = table_bac,
                metadata= metadata_bacteria,
                formula_str = "Type_of_soil*Treatment",
                method = "compositional",
                test = "permanova",
                permutations = 999)
```

```{r, fig.width=8, fig.height=4}
beta_test_table(table = table_fung,
                metadata= metadata_fungi,
                formula_str = "Type_of_soil*Treatment",
                method = "compositional",
                test = "permanova",
                permutations = 999)
```

### 💻 Beta diversity partitioning
Beta diversity can also be decomposed into different components. The `beta_partition_ord_plot()` function partitions beta diversity into the turnover and nestedness components based on selected dissimilarity indices.

The `index` argument specifies the dissimilarity family used for the partition (e.g., Jaccard or Sørensen). The `group_col` and `shape_col` parameters define how samples are visualized according to metadata variables.

```{r, fig.width=14, fig.height=5.5}
beta_partition_ord_plot(
  table = table_bac,
  metadata = metadata_bacteria,
  index = "jaccard",
  group_col = "Type_of_soil",
  shape_col = "Treatment", 
  save_table = FALSE)
```

```{r, fig.width=14, fig.height=5.5}
beta_partition_ord_plot(
  table = table_fung,
  metadata = metadata_fungi,
  index = "jaccard",
  group_col = "Type_of_soil",
  shape_col = "Treatment", 
  save_table = FALSE)
```

### 🖥️ Pairwise beta diversity visualization
The `beta_dissimilarity_plot()` function summarizes pairwise beta diversity values using boxplots, allowing comparison of diversity patterns among groups of samples. This funciont uses betapart on incidence data using Sorensen or Jaccard index: shared, turnover or nestedness. 

The `condition1_col` and `condition2_col` parameters specify metadata variables used to define the groups being compared. The `partition` argument determines which component of beta diversity is displayed (e.g., shared or unique), while `family` defines the dissimilarity index used.

By default `beta_dissimilarity_plot()` keeps every pairwise combination possible of `condition1_col`, including a group compared against itself (e.g. `"Roots_vs_Roots"`). The `comparison_condition1` argument restricts the plot to specific `"value1_vs_value2"` labels (the two values sorted alphabetically and joined with `"*_vs_*"`). Here we build that list programmatically to keep only the comparisons of every other compartment **against Roots**, dropping self-comparisons:

```{r}
groups_bac <- setdiff(unique(as.character(metadata_bacteria$Type_of_soil)), "Roots")
vs_roots_bac <- paste0(pmin(groups_bac, "Roots"), "_vs_", pmax(groups_bac, "Roots"))
vs_roots_bac
```

As with the alpha diversity plots, `x_label_angle` rotates the x-axis tick labels (default `0`), `strip_text_bold` toggles bold facet strip titles (default `FALSE`), and `aspect_ratio` fixes the panel's height/width ratio. As with the alpha plots, `stat` (e.g. `"wilcox.test"`, `"kruskal.test"`, `"anova"`) adds a statistical comparison across the boxplots, passed through to `ggpubr::stat_compare_means()`.

If `save_table = TRUE`, the function also returns a table containing the beta diversity values used to generate the plot.

```{r, fig.width=9, fig.height=6}
beta_dissimilarity_plot(table = table_bac,
                       metadata = metadata_bacteria,
                       comparison_condition1 = vs_roots_bac,
                       condition1_col = "Type_of_soil",
                       condition2_col = "Treatment",
                       x_axis_title = "Samples",
                       x_label_angle = 45,
                       stat = "kruskal.test",
                       partition = "shared", 
                       family = "jaccard",
                       save_table = FALSE)

```

```{r}
groups_fung <- setdiff(unique(as.character(metadata_fungi$Type_of_soil)), "Roots")
vs_roots_fung <- paste0(pmin(groups_fung, "Roots"), "_vs_", pmax(groups_fung, "Roots"))
```

```{r, fig.width=9, fig.height=6}
beta_dissimilarity_plot(table = table_fung,
                       metadata = metadata_fungi,
                       comparison_condition1 = vs_roots_fung,
                       condition1_col = "Type_of_soil",
                       condition2_col = "Treatment",
                       x_axis_title = "Samples",
                       x_label_angle = 45,
                       stat = "kruskal.test",
                       partition = "shared",
                       family = "jaccard",
                       save_table = FALSE)

```

### 🖥️ Pairwise turnover: between-group vs. within-group
`beta_turnover_plot()` reports the ASV turnover component of Hill-number-based beta diversity (*q*=0, 1, 2), one facet row per Hill order.

`beta_turnover_plot()` answers a different question than `beta_dissimilarity_plot()`: for **one reference group** (e.g. Rhizosphere), is its turnover **against another group** (e.g. Roots, a between-group comparison) greater than its turnover **against itself** (Rhizosphere-vs-Rhizosphere pairs, a within-group baseline)? If between-group turnover is significantly higher, that's evidence the two compartments really do host distinct communities, beyond the variability already present among replicates of the same compartment.

So the plot below has one purple (this color can be set) strip on top labeled **"Rhizosphere"**, the fixed reference group in both comparisons and,inside it, two boxes: **"Roots"** (Rhizosphere vs. Roots, between-group) and **"Rhizosphere"** (Rhizosphere vs. Rhizosphere, within-group). The strip and the "Rhizosphere" box both being labeled the same is intentional, not a duplicate, the strip names the reference, the box names what it's being compared *against* (itself, in that one case).

`comparison_condition1` selects which `"value1_vs_value2"` combinations to keep always with the reference group (`Rhizosphere`) first. `comparison_condition2` keeps only the same treatment pairs, so turnover isn't confounded by comparing samples from different treatments.

`color_facets_x` colors the top facet strip (`condition1.x`) and `color_axis_x` colors the two boxes (`condition1.y`). Tick labels are hidden by default since the groups are already identified in the legend. Set `show_x_labels = TRUE` to bring them back. `stat` adds a statistical comparison between the two boxes.

```{r, warning=FALSE, fig.width=3.5, fig.height=4.5, out.width="60%"}
beta_turnover_plot(
  table                 = table_fungi,
  metadata              = metadata_fungi,
  comparison_condition1 = c("Rhizosphere_vs_Roots", "Rhizosphere_vs_Rhizosphere"),
  comparison_condition2 = c("TC_vs_TC", "TD_vs_TD", "TED_vs_TED"),
  condition1.x          = "Type_of_soil.x",
  condition1.y          = "Type_of_soil.y",
  condition2.x          = "Treatment.x",
  condition2.y          = "Treatment.y",
  color_facets_x        = "#5D478B",
  color_axis_x          = c("Roots" = "#56B4E9", "Rhizosphere" = "#E69F00"),
  stat                  = "wilcox.test"
)$plot
```

### 💻 Distance-decay of community similarity
The function `beta_decay_plot()` tests whether communities that are geographically closer are also more similar in composition. It computes pairwise community dissimilarity (`distance`, e.g. `"jaccard"`, `"bray"`, `"horn"`) and pairwise geographic distances from sample coordinates, runs a Mantel test between them, and plots similarity (1 − dissimilarity) against distance with a fitted regression line.

`lat_col`/`lon_col` name the latitude/longitude columns in `metadata` (decimal degrees), and the optional `group_col` runs a separate Mantel test and regression line per group instead of one global test.

We construct example data for this part:

```{r, warning=FALSE, fig.width=8, fig.height=6}
loc_coords <- data.frame(
  Loc = 0:8,
  lat = 19.0 + seq(0, 0.8, length.out = 9),
  lon = -99.0 + seq(0, 0.8, length.out = 9)
)
metadata_bacteria$lat <- loc_coords$lat[match(metadata_bacteria$Loc, loc_coords$Loc)]
metadata_bacteria$lon <- loc_coords$lon[match(metadata_bacteria$Loc, loc_coords$Loc)]

#filtering uncultivated
metadata_bacteria_decay <- metadata_bacteria %>%
  filter(Type_of_soil != "Uncultivated")
```


```{r, warning=FALSE, fig.width=8, fig.height=6}
beta_decay_plot(
  table     = table_bac,
  metadata  = metadata_bacteria_decay,
  lat_col   = "lat",
  lon_col   = "lon",
  distance  = "jaccard",
  group_col = "Type_of_soil"
)
```

## 📊 Differential abundant analysis
`MicroBioMeta` has several functions to explore **differentially abundant taxa**. These analyses help identify microbial features that vary significantly across experimental groups.

Two functions rely on a **compositional data analysis framework implemented in the package** [ALDEx2](https://www.bioconductor.org/packages/release/bioc/html/ALDEx2.html). This approach accounts for the compositional nature of sequencing data by using Monte Carlo Dirichlet instances and centered log-ratio transformations and in the package [ANCOMBC2](https://bioconductor.org/packages/ANCOMBC/) an alternative differential abundance method that explicitly models compositional bias and structural zeros.

The functions:

-   `aldex_volcano_plot()` visualizes differential abundance results using a **volcano plot**.

-   `aldex_heatmap_plot()` visualizes significant taxa in a **heatmap**.

-   `ancombc_plot()` returns a bar plot (when 2 groups) and a heatmap (when 3+ groups).


Additionally, two complementary approaches help identify **important taxa that discriminate between groups**:

-   `ratios_bubble_plot()` visualizes taxa based on **log-ratio differences between conditions**.

-   `random_forest_lollipop_plot()` uses a **Random Forest model** to identify features that best predict a given variable or condition, visualized as a **lollipop plot of feature importance**.



### 📊 Example: Bacterial data
First, we filter the metadata to retain only two soil compartments to apply ALDEx2.

```{r}
metadata_bacteria_compar <-  metadata_bacteria %>% 
  filter(Type_of_soil == "Roots" | Type_of_soil =="Rhizosphere")

table_bacteria_compar <- table_bacteria[match(metadata_bacteria_compar$SAMPLEID, colnames(table_bacteria))]

table_bac_compar <- merge_feature_taxonomy(table_bacteria_compar, taxonomy_bacteria)

```

We can visualize differential abundance using a **volcano plot**. Significant taxa are labeled with their name; `label_size` controls the font size of these labels, and `filter_uncultured = TRUE` drops "uncultured"/"unculture" taxa from the labels (they are still plotted as points, just not labeled).

```{r, fig.width=7, fig.height=5.5, message=FALSE, warning=FALSE}
aldex_volcano_plot(table = table_bac_compar,
                   metadata = metadata_bacteria_compar,
                   col_cond = "Type_of_soil",
                   type = "volcano",
                   label_size = 3,
                   filter_uncultured = TRUE )

```

### 📊 ANCOMBC2 as an alternative method
`ancombc_plot()` provides an alternative to the ALDEx2-based functions above. This functions depends on the Bioconductor packages `ANCOMBC` and `phyloseq`:

```{r, eval=FALSE}
# install.packages("BiocManager")
BiocManager::install(c("ANCOMBC", "phyloseq"))
```

`tax_level` groups the table to a taxonomic rank before testing (default `"Genus"`), and `p_adj_method`/`prv_cut` control the multiple-testing correction and prevalence filter.

⚠️ With a low `prv_cut`, rare/sparse taxa can trigger a `"Zero variances have been detected..."` error from ANCOMBC2's internal bias-correction step on some runs (it doesn't always happen, it depends on which taxa the bootstrap iteration happens to keep). Raising `prv_cut` filters those sparse taxa out before testing and avoids it.

```{r, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
ancombc_plot(
  table        = table_bac_compar,
  metadata     = metadata_bacteria_compar,
  col_cond     = "Type_of_soil",
  tax_level    = "Genus",
  prv_cut      = 0.3,
  p_adj_method = "BH"
)
```

### 📊 Example: Fungal data
We follow the same procedure but visualize the results with a **heatmap**. This funtcions uses `ComplexHeatmap`. 

```{r}
metadata_fungi_compar <-  metadata_fungi %>%
  filter(Type_of_soil == "Bulk soil" | Type_of_soil =="Rhizosphere")

table_fungi_compar <- table_fung[match(metadata_fungi_compar$SAMPLEID, colnames(table_fung))]

table_fung_compar <- merge_feature_taxonomy(table_fungi_compar, taxonomy_fungi)

```

```{r, fig.width=9, fig.height=4, message=FALSE, warning=FALSE}
aldex_heatmap_plot(table = table_fung_compar,
                   metadata = metadata_fungi_compar,
                   col_cond = "Type_of_soil")

```

### 📊 Identifying important taxa
The functions `random_forest_lollipop_plot()` and `ratios_bubble_plot()` can also be used to highlight taxa that contribute most strongly to the observed differences between conditions.

For example, a **Random Forest model** can be used to identify the taxa that best predict the soil compartment:

```{r, fig.width=7, fig.height=6}
random_forest_lollipop_plot(
  table = table_bac, 
  metadata = metadata_bacteria, 
  top_n = 10, 
  variable_to_predict = "Type_of_soil")
```

Alternatively, taxa with the strongest **log-ratio differences between two conditions** can be visualized using:

```{r, fig.width=8, fig.height=9}
ratios_bubble_plot(table = table_fung_compar,
           metadata = metadata_fungi_compar,
           condition_col = "Type_of_soil",
           top_n = 20,
           condition_A = "Rhizosphere",
           condition_B = "Bulk soil")
```

## 💻 Environmental analyses
`MicroBioMeta` provides two functions to explore the relationship between **environmental variables** and **microbial community composition**:

-   `corr_env_abund_plot()` — computes correlations between taxonomic abundance and environmental variables.

-   `cca_rda_biplot()` — performs constrained ordination using **Canonical Correspondence Analysis (CCA)** or **Redundancy Analysis (RDA)**.

#### 🖥️ Preparing the environmental table
Environmental variables must be organized in a **data frame where rows correspond to samples and columns correspond to environmental variables**.

```{r}
env_table_fung <- metadata_fungi %>%
  dplyr::select(SAMPLEID, pH:Arbus_per) %>%
  remove_rownames() %>%
  column_to_rownames(var = "SAMPLEID")
```

#### 🖥️ Correlation between environmental variables and taxonomic abundance
The function `corr_env_abund_plot()` calculates correlations between **environmental variables** and **taxonomic relative abundances** at a selected taxonomic level and returns a **correlation heatmap** showing how environmental variables relate to taxonomic abundances.

Example using **phylum-level abundances**:

```{r, fig.width=9, fig.height=8}
corr_env_abund_plot(table = table_fung,
                    env_table = env_table_fung,
                    metadata = metadata_fungi,
                    level = "phylum",
                    save_table = FALSE)

```



### 🖥️ Alternative visualization
Correlations can also be displayed using circle markers instead of tiles:

```{r, fig.width=9, fig.height=8}
corr_env_abund_plot(table = table_fung,
                    env_table = env_table_fung,
                    metadata = metadata_fungi,
                    level = "phylum",
                    save_table = FALSE,
                    geom = "circle")

```

#### 🖥️ Constrained ordination (CCA / RDA)
The function `cca_rda_biplot()` performs a constrained ordination analysis to evaluate how environmental variables explain variation in microbial community composition.

Before running the analysis, ensure that the **sample order in the environmental table matches the metadata**:

```{r}
env_table_fung <- env_table_fung[
  match(metadata_fungi$SAMPLEID, rownames(env_table_fung)), , drop = FALSE]

```

`scale_arrows` is a parameter that helps you to visualize better the environmental vectors:

```{r, fig.width=8, fig.height=6}
cca_rda_biplot(table = table_fung,
               env_data = env_table_fung,
               metadata = metadata_fungi,
               env_vars = c("pH","TN","WHC", "EC", "Clay"),
               analysis = "RDA",
               show_all_env_vectors = TRUE,
               group_col = "Type_of_soil",
               scale_arrows = 3)

```

This analysis produces a **biplot** where:

-   **Points** represent samples

-   **Colors** indicate sample groups defined in the metadata.

-   **Arrows** represent environmental variables and their direction of influence on community structure.

-   Environmental vectors pointing in similar directions indicate **positive associations**, while vectors pointing in opposite directions suggest **negative relationships**


## Session info

```{r}
sessionInfo()
```
