---
title: "Metagenomic"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Metagenomic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

🍄 This tutorial shows how to use `MicroBioMeta` on **shotgun metagenomic** data, as an alternative to the amplicon (metabarcoding) workflow covered in the [Metabarcoding vignette](Metabarcoding.html). The functions and most of their parameters are exactly the same across both workflows, so they aren't re-explained here — only what's specific to metagenomic data is.

For this example, reads were taxonomically classified with [`Kraken2`](https://ccb.jhu.edu/software/kraken2/) and abundance-corrected at the species level with [`Bracken`](https://ccb.jhu.edu/software/bracken/).

The data comes from forest soils along the Iztaccíhuatl-Popocatépetl (PNIP) and La Malinche (PNML) transects, Central Mexico:

[Hereira-Pacheco, S. E. et al. (2025). Metagenomic analysis of the diversity and composition of the fungal communities of forest soils of the transect Iztaccíhuatl-Popocatépetl and La Malinche. *PeerJ*, 13, e18323.](https://peerj.com/articles/18323/)

Raw data, metadata and the original analysis scripts: [github.com/Steph0522/Fungal_Metagenomes_PNIP_PNML](https://github.com/Steph0522/Fungal_Metagenomes_PNIP_PNML)

## Loading data 💻

First, let's load `MicroBioMeta` and `tidyverse` (used for data wrangling):

```{r "start", message=FALSE}
library("MicroBioMeta")
library(tidyverse)
```

Unlike the QIIME2 workflow in the metabarcoding vignette, there's no `.qza` import step here — Kraken2/Bracken output is plain text. Each sample was processed separately, so `bracken_species` reports were merged into a single table beforehand (one column per sample, named `kraken_pluspfp_<id_metagenome>_report_bracken_species`) with a **`taxonomy`** column already attached as the last column:

```{r}
table_kraken <- read.delim("articles/data/table_kraken.txt", row.names = 1, check.names = FALSE)

table_kraken[1:3, c(1:2, ncol(table_kraken))]
```

Because taxonomy already ships with the table, `merge_feature_taxonomy()` isn't needed for this dataset — it's only required when counts and taxonomy come as two separate objects, as in the metabarcoding example.

⚠️ Taxonomy strings from Kraken2 use a `k__`/`p__`/`c__`/... prefix like SILVA/GTDB, but ranks are separated by `"; "` (semicolon **and space**) instead of `";"`. `MicroBioMeta` already accounts for this wherever a `taxonomy_db` argument exists — just pass `taxonomy_db = "Kraken2"`.

Now the metadata. These sites were sampled in **two seasons** (dry/*secas* in March and rainy/*lluvias* in September), but the metagenomes were sequenced only once per site — from the **dry-season** sampling — so that's the metadata to use here; the rainy-season file records the same sites' soil chemistry at a different time of year and doesn't correspond to a separate set of metagenomes.

```{r}
metadata_meta <- read.delim("articles/data/metadata_metagenomic.txt", check.names = FALSE)

head(metadata_meta[, 1:10])
```

Kraken column names encode a `id_metagenome` number (e.g. `kraken_pluspfp_100_report_bracken_species`) rather than the sample's actual ID, so we recover it with a regular expression and use `metadata_meta$id_metagenome` to rename each column to its `SampleID`:

```{r}
sample_cols <- setdiff(colnames(table_kraken), "taxonomy")

id_metagenome <- as.integer(gsub("kraken_pluspfp_|_report_bracken_species", "", sample_cols))

colnames(table_kraken)[match(sample_cols, colnames(table_kraken))] <-
  metadata_meta$SAMPLEID[match(id_metagenome, metadata_meta$id_metagenome)]
```

Two metadata columns we'll use repeatedly below as grouping variables — `Poligono` (6 sampling blocks, 6 replicates each) and `Sitio` (2 sub-sites per block) — come in as plain numbers. `Sitio` just needs to be character so it's treated as a group rather than a continuous covariate; `Poligono` is relabeled to `"Pol1"`...`"Pol6"` with explicit, numerically-sorted factor levels, so every plot below lists/facets the blocks in that order instead of whatever order they happen to first appear in the data:

```{r}
metadata_meta$Poligono <- factor(paste0("Pol", metadata_meta$Poligono), levels = paste0("Pol", 1:6))
metadata_meta$Sitio <- as.character(metadata_meta$Sitio)
```

`table_meta`'s sample columns are also reordered to match `metadata_meta`'s row order (rather than the ascending `id_metagenome` order they came in), for the same reason — several plotting functions derive their group order from column/row appearance order:

```{r}
table_meta <- table_kraken[, c(metadata_meta$SAMPLEID, "taxonomy")]
```

`table_meta` is now in the same `taxonomy`-as-last-column format used throughout the package, ready for every `MicroBioMeta` function — equivalent to `table_bac`/`table_fung` in the metabarcoding vignette.

## Composition exploration 📊

### Abundance barplot 📊

Same function as in the metabarcoding vignette; the only difference is `taxonomy_db = "Kraken2"`, which tells the taxonomy-parsing logic to expect Kraken2/Bracken-style strings.

```{r, fig.width=9, fig.height=6}
abundance_bar_plot(table = table_meta,
                  metadata = metadata_meta,
                  taxonomy_db = "Kraken2",
                  level = "genus",
                  top_n = 20,
                  x_col = "Poligono",
                  add_remained = TRUE,
                  label = "Genus",
                  save_table = FALSE)
```

### Abundance heatmap 🖥️

```{r, warning=FALSE, fig.width=8, fig.height=8.5}
abundance_heatmap_plot(table = table_meta,
                  metadata = metadata_meta,
                  top_n = 20,
                  show_column_names = FALSE,
                  condition1 = "Poligono")
```

## Alpha diversity 📊

### Correlation among Hill numbers 📈

```{r, warning=FALSE, fig.width=12, fig.height=4.5}
alpha_hill_corr_plot(table = table_meta)
```

### Alpha diversity visualization 📊

```{r, warning=FALSE, fig.width=9, fig.height=5}
alpha_hill_plot(table = table_meta,
                metadata = metadata_meta,
                x_col = "Poligono",
                fill_col = "Poligono",
                facet_by = "Sitio",
                facet_orientation = "horizontal",
                save_table = FALSE)
```

Richness (*q*=0) is nearly saturated and flat across blocks here — this genus-level Bracken table simply doesn't have much room left to vary at *q*=0 — while *q*=1 and *q*=2 (which weight by relative abundance) do pick up differences between blocks.

### Alpha diversity along a continuous gradient

```{r, warning=FALSE, fig.height=4, fig.width=10}
alpha_decay_plot(table = table_meta,
                 metadata = metadata_meta,
                 cont_var = "pH",
                 x_axis_title = "Soil pH")
```

Diversity at *q*=1 and *q*=2 increases significantly with soil pH; *q*=0 doesn't, for the same ceiling-effect reason as above.

## Beta diversity

### Ordination of community composition

The CLR-based `distance = "compositional"` (used for PCA in the metabarcoding vignette) is dominated here by a single compositionally extreme sample, which stretches the plot and squashes everything else into one corner. Bray-Curtis dissimilarity with PCoA is far less sensitive to that one sample, so it's used instead — this is also closer to what the original study behind this dataset used for its own ordination.

```{r, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
beta_ord_plot(table = table_meta,
              metadata = metadata_meta,
              distance = "bray",
              ordination = "PCoA",
              group_col = "Poligono")
```

### Statistical tests for beta diversity

`method = "bray"` matches the `distance = "bray"` used for the PCoA above, so the ordination and the significance test are evaluating the same notion of dissimilarity. `Poligono` comes out strongly significant (R² = 0.47).

```{r, fig.width=8, fig.height=4}
beta_test_table(table = table_meta,
                metadata = metadata_meta,
                formula_str = "Poligono*Sitio",
                method = "bray",
                test = "permanova",
                permutations = 999)
```

### Beta diversity partitioning

`beta_partition_ord_plot()` works the same way here as in the metabarcoding vignette — see that vignette for a worked example. It's skipped for *this particular* dataset because the nestedness component of the Jaccard/Sørensen partition is close to singular (turnover dominates almost completely over nestedness among these forest-soil fungal communities), which makes the underlying ordination fail; that's a property of this community's structure, not something specific to metagenomic data.

## Differential abundant analysis

For the heatmap and ANCOMBC2 versions we need a 2-group subset of the 6-level `Poligono`, the same way the metabarcoding vignette filters `Type_of_soil` down to two levels — blocks `2` and `5` are used here as they show a clear compositional difference:

```{r}
metadata_meta_compar <- metadata_meta %>%
  filter(Poligono == "Pol2" | Poligono == "Pol5")

table_meta_compar <- table_meta[, c(match(metadata_meta_compar$SAMPLEID, colnames(table_meta)), ncol(table_meta))]
```

```{r, fig.width=7, fig.height=6, message=FALSE, warning=FALSE}
aldex_heatmap_plot(table = table_meta_compar,
                   metadata = metadata_meta_compar,
                   col_cond = "Poligono")
```

```{r, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
ancombc_plot(
  table        = table_meta_compar,
  metadata     = metadata_meta_compar,
  col_cond     = "Poligono",
  tax_level    = "Genus",
  prv_cut      = 0.05,
  p_adj_method = "BH"
)
```

### Identifying important taxa

```{r, fig.width=7, fig.height=6}
random_forest_lollipop_plot(table = table_meta, metadata = metadata_meta, top_n = 10, variable_to_predict = "Poligono")
```

```{r, fig.width=8, fig.height=9}
ratios_bubble_plot(table = table_meta_compar,
           metadata = metadata_meta_compar,
           condition_col = "Poligono",
           top_n = 20,
           condition_A = "Pol2",
           condition_B = "Pol5")
```

## Environmental analyses

The metadata already carries the soil physicochemical variables measured for these samples, so the environmental table is built the same way as in the metabarcoding vignette:

```{r}
env_table_meta <- metadata_meta %>%
  dplyr::select(SAMPLEID, pH:ARENA) %>%
  remove_rownames() %>%
  column_to_rownames(var = "SAMPLEID")
```

### Correlation between environmental variables and taxonomic abundance

```{r, fig.width=9, fig.height=3.2}
corr_env_abund_plot(table = table_meta,
                    env_table = env_table_meta,
                    metadata = metadata_meta,
                    taxonomy_db = "Kraken2",
                    level = "phylum",
                    save_table = FALSE)
```

### Constrained ordination (CCA)

```{r}
env_table_meta <- env_table_meta[
  match(metadata_meta$SAMPLEID, rownames(env_table_meta)), , drop = FALSE]
```

```{r, fig.width=8, fig.height=6}
cca_rda_biplot(table = table_meta,
               env_data = env_table_meta,
               metadata = metadata_meta,
               env_vars = c("pH", "MO", "N", "P", "K"),
               analysis = "CCA",
               show_all_env_vectors = TRUE,
               group_col = "Poligono",
               scale_arrows = 3)
```
