---
title: "metabarcoding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{metabarcoding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

For this example, we are going to use metabarcoding data: 16S for bacteria and 18S for fungi.

The data is taken from: [16S](https://www.nature.com/articles/s41598-021-81551-7) [18S](https://www.sciencedirect.com/science/article/abs/pii/S1754504823000028)

First, Let's call the `MicroBioMeta` package:

```{r "start", message=FALSE}
library("MicroBioMeta")

```

Let's call other packages needed:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(qiime2R)
```

tidyverse is used for all the data wrangling and visualizations

Then, Load the data and reformat:

```{r}

bacteria_table <- qiime2R::read_qza("../man/Data/dada2.200.trimmed.single.filtered.cleaned.qza")$data %>% as.data.frame()

fungi_table <- qiime2R::read_qza("../man/Data/merge_table_240_noplant_filtered_nous.qza")$data %>% 
  as.data.frame()

metadata <- read.delim("../man/Data/FINALMAP.txt", check.names = F) %>% rename(SAMPLEID =`#SampleID`) %>% 
  filter(Month == "2") %>%
  mutate(Treatment = factor(
    case_when(
      Treatment == 1 ~ "TC",
      Treatment == 2 ~ "TD",
      Treatment == 3 ~ "TED",
      TRUE ~ as.character(Treatment)
    ),
    levels = c("TC", "TD", "TED")
  ))

metadata2 <- read.delim("../man/Data/FINALMAP18S.txt", check.names = F)%>% rename(SAMPLEID=`#SampleID`)%>%
  mutate(
    Treatment = factor(
      case_when(
        Treatment == 1 ~ "TC",
        Treatment == 2 ~ "TD",
        Treatment == 3 ~ "TED",
        TRUE ~ as.character(Treatment)
      ),
      levels = c("TC", "TD", "TED")
    )
  )



bacteria_table_filter <- bacteria_table[match(metadata$SAMPLEID, colnames(bacteria_table)), ]


taxonomy_bacteria <- read_qza("../man/Data/taxonomy_dada2.200.trimmed.single.seqs.qza")$data %>%
  as.data.frame() %>% rename(taxonomy = Taxon) %>% 
  dplyr::select(-Confidence) %>% 
  column_to_rownames(var = "Feature.ID")

taxonomy_fungi <- read_qza("../man/Data/taxonomy_blast_240_0.97.qza")$data %>% as.data.frame() %>% 
  rename(taxonomy = Taxon) %>% 
  dplyr::select(-Consensus) %>% 
  column_to_rownames(var = "Feature.ID")


```

## Pre-processing data

In order to use the data as input for functions of `MicroBioMeta` should contain a biom format with the column of *taxonomy* at the end.

```{r}

table_bac <- merge_feature_taxonomy(table = bacteria_table_filter,
                                    taxonomy = taxonomy_bacteria)

table_fung <- merge_feature_taxonomy(table = fungi_table, 
                                     taxonomy = taxonomy_fungi)

```

## Composition exploration

### Abundance barplots

```{r}
abundance_barplot(table = table_bac,
                  metadata = metadata,
                  taxonomy_db = "silva",
                  level = "phylum", 
                  x_col = "Type_of_soil", 
                  facet_col = "Treatment", 
                  add_remained = TRUE,
                  label =   "Phylum",
                  save_table = FALSE)+
  theme(axis.text.x = element_text(angle = 50, hjust = 0.95))

```

```{r}
abundance_barplot(table = table_fung,
                  metadata = metadata2,
                  x_col = "Treatment", 
                  facet_col = "Type_of_soil", 
                  add_remained = TRUE,  
                  label =   "Genera",
                  save_table = FALSE)+
  theme(axis.text.x = element_text(angle = 50, hjust = 0.95))
```

### Abundance heatmaps

```{r, warning=FALSE}
abundance_heatmap_plot(table = table_bac,
                  metadata = metadata,
                  top_n = 30,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil", 
                  condition2 = "Treatment")
```

```{r, warning=FALSE}
abundance_heatmap_plot(table = table_fung,
                  metadata = metadata2,
                  top_n = 30,
                  show_column_names = FALSE,
                  condition1 = "Type_of_soil", 
                  condition2 = "Treatment")
```

## Alpha diversity

```{r, warning=FALSE}
alpha_hill_corrplot(table = table_bac)

```

```{r, warning=FALSE}
alpha_hill_corrplot(table = table_fung)
```

```{r, warning=FALSE}
alpha_hill_plot(table = table_bac, 
                metadata = metadata%>% filter(SAMPLEID %in% colnames(table_bac)),
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "vertical",
                save_table = FALSE)
```

```{r, warning=FALSE}
alpha_hill_plot(table = table_fung, 
                metadata = metadata%>% filter(SAMPLEID %in% colnames(table_fung)),
                x_col = "Treatment",
                fill_col = "Treatment",
                facet_by = "Type_of_soil",
                facet_orientation = "horizontal",
                save_table = FALSE)
```

## Beta diversity

```{r, warning=FALSE}
beta_div_plot(table = table_bac, 
              metadata = metadata, 
              group_col = "Type_of_soil",
              shape_col = "Treatment")
```

```{r, warning=FALSE}
beta_div_plot(table = table_fung, 
              metadata = metadata2, 
              group_col = "Type_of_soil",
              shape_col = "Treatment",distance = "aitchison", ordination = "NMDS")
```
