---
title: "Installation"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

`MicroBioMeta` requires **R >= 4.3**.

## Installing MicroBioMeta

While the package is in review for Bioconductor, install it from GitHub:

```{r}
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("Steph0522/MicroBioMeta")
```

Once accepted, it will also be installable the standard Bioconductor way:

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("MicroBioMeta")
```

## Dependencies

Every package `MicroBioMeta` needs for its core functions (CRAN and Bioconductor packages: `vegan`, `hillR`, `betapart`, `ALDEx2`, `ANCOMBC`, `phyloseq`, `ComplexHeatmap`, `ggplot2`, and others) installs automatically with either method above — no extra steps needed for regular use.

## Additional packages for the tutorials

A couple of examples in the **Tutorials** use packages that aren't required by the package itself, only by that specific example:

- **[`qiime2R`](https://github.com/jbisanz/qiime2R)** (GitHub-only) — used in the Metabarcoding/Amplicón tutorial to import `.qza` artifacts exported from QIIME2:

  ```{r}
  devtools::install_github("jbisanz/qiime2R")
  ```

  This is only needed if your own data comes from QIIME2 as `.qza` files. If you already have plain `.txt`/`.csv` tables, you can skip it entirely and read those directly with `read.delim()`/`read.csv()`.

- **[`networkD3`](https://CRAN.R-project.org/package=networkD3)** (CRAN) — used by `abundance_sankey_plot()` for the interactive Sankey diagram:

  ```{r}
  install.packages("networkD3")
  ```

## Troubleshooting Bioconductor packages

If `BiocManager::install()` isn't already set up, run this once first:

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(version = "3.19")
```
